# Aquaculture

**Description**: Aquaculture indicators

**Found in**: State of the Ecosystem - Gulf of Maine & Georges Bank (2021), State of the Ecosystem - Mid-Atlantic (2021)

**Indicator category**: Synthesis of published information

**Contributor(s)**: Christopher Schillaci, Maine DMR, NH DES, MA DMF, RI CRMC, MD DNR
  
**Data steward**: Chris Schillaci <christopher.schillaci@noaa.gov>
  
**Point of contact**: Chris Schillaci <christopher.schillaci@noaa.gov>
  
**Public availability statement**: Source data are publicly available

## Impact Summary

Overall oyster production is increasing and production per area is also improving.  Aquaculture reporting is state specific and because of this can be difficult to collate in a meaningful way. However, this is improving year to year as aquaculture increases across the coast. The data included here only holds oyster numbers at this time. 

### Plotted data 
Code for plotting data included in the State of the Ecosystem report can be found [here](https://github.com/NOAA-EDAB/ecodata/blob/master/chunk-scripts/human_dimensions.Rmd-oyster-aqua.R). 

```{r, code = readLines("https://raw.githubusercontent.com/NOAA-EDAB/ecodata/master/chunk-scripts/human_dimensions_NE.Rmd-aquaculture-pieces.R"), fig.cap=".",echo = F, eval =T, warning = F, message=F}
```

```{r, code = readLines("https://raw.githubusercontent.com/NOAA-EDAB/ecodata/master/chunk-scripts/human_dimensions_NE.Rmd-aquaculture-pa.R"), fig.cap=".",echo = F, eval =T, warning = F, message=F}
```

### Raw data 
The raw data is available in the [`ecodata` package](https://noaa-edab.github.io/ecodata/landing_page). 

```{r}
library(DT)
datatable(ecodata::aquaculture, options = list(pageLength = 10))
``` 

## Methods
### Data Sources 
Data was synthesized from state specific sources, listed below. 

* [State of Maine, Department of Marine Resources.](https://www.maine.gov/dmr/aquaculture/data/index.html)                

* [State of New Hampshire, Marine Aquaculture Compendium](https://drive.google.com/file/d/1eCg0cP2rsjZ0AAloPuxIyDiA01urOcjR/view?usp=sharing)

* [State of Massachusetts, Division of Marine Fisheries](https://www.mass.gov/service-details/dmf-annual-reports)

* [State of Rhode Island, Coastal Resource Management Council](http://www.crmc.ri.gov/aquaculture.html)        

* [State of Maryland, Aquaculture Coordinating Council](https://calendarmedia.blob.core.windows.net/assets/1495a281-9eab-422a-9f90-a16ac9686db8.pdf) 

### Data Extraction
Production data was pulled from the state sources above. 

### Data analysis
Divsion of production in pieces by shellfish lease area acres to calculate production per acre.

### Data processing

Aquaculture data were formatted for inclusion in the `ecodata` R package using the code found [here](https://github.com/NOAA-EDAB/ecodata/blob/master/data-raw/get_aquaculture.R).

