# Aquaculture Production {#aquaculture}

**Description**: Oyster production: number of oysters harvested from aquaculture.

**Indicator family**: 

- [X] Megafauna
- [X] Social
- [X] Economic


**Contributor(s)**: Christopher Schillaci, Maine DMR, NH DES, MA DMF, RI CRMC, MD DNR

**Affiliations**: NOAA

```{r echo=FALSE}
knitr::opts_chunk$set(echo = F)
library(ecodata)
```
## Introduction to Indicator
Aquaculture can produce seafood as well as other products. Here we summarize available data on aquaculture production from the Northeast US.

## Key Results and Visualizations
Overall, oyster production has generally increased and production per area has also improved. Aquaculture reporting is state specific, which can make it difficult to collate in a meaningful way. However, reporting coverage is improving year to year as aquaculture increases across the coast. The data included here only include oyster numbers at this time. Data after 2017 have decreased but, due to a lack of data from recent years, it is unclear if this has continued.

```{r plot_aquacultureMidAtlantic}
# Plot indicator
ggplotObject <- ecodata::plot_aquaculture(report = 'MidAtlantic', n = 10)
ggplotObject
```

```{r plot_aquacultureNewEngland}
# Plot indicator
ggplotObject <- ecodata::plot_aquaculture(report = 'NewEngland', n = 10)
ggplotObject
```


## Indicator statistics 
Spatial scale: Mid Atlantic and New England

Temporal scale: Annual

**Synthesis Theme**:

- [X] Multiple System Drivers


```{r autostats_aquaculture}
# Either from Contributor or ecodata
```

## Implications
Aquaculture production contributes to overall seafood production in the Northeast US. This indicator provides only a portion of aquaculture production in the region, but increases in this portion are apparent across the whole time series. Recent decreases (since 2017) may indicate a downward change in this aquaculture portion; however, there is not enough data from years following 2017 to indicate any significant decreasing trends.

## Get the data

**Point of contact**: [Chris Schillaci <christopher.schillaci@noaa.gov>](mailto:Chris Schillaci <christopher.schillaci@noaa.gov>){.email}

**ecodata name**: `ecodata::aquaculture`

**Variable definitions**

Pieces: number of oysters produced (all regions)                 Shellfish lease Acres: area used for shellfish production (New England states only), acres  
Production/Acre: Pieces divided by Shellfish lease acres (New England states only)

```{r vars_aquaculture}
# Pull all var names
vars <- ecodata::aquaculture |>
   dplyr::select(Var) |>
   dplyr::distinct()

DT::datatable(vars)
```
**Indicator Category**:

- [X] Syntheses of published information


## Public Availability

Source data are publicly available.

## Accessibility and Constraints

_No response_

**tech-doc link**
<https://noaa-edab.github.io/tech-doc/aquaculture.html>

