# Aquaculture Production {#aquaculture}

```{r echo=FALSE}
knitr::opts_chunk$set(echo = F)
library(ecodata)
library(downloadthis)
```

<a href="https://noaa-edab.github.io/catalog/page_status.html#under-review" class="btn btn-danger button_small"><i class="fa fa-warning"></i> Under Review</a>

*Last updated July 23, 2026*

**Description**: Oyster production: number of oysters harvested from aquaculture.

**Contributor(s)**: Christopher Schillaci, Maine DMR, NH DES, MA DMF, RI CRMC, MD DNR

**Affiliations**: NOAA

**Indicator Family**: 

- [X] Megafauna
- [X] Social
- [X] Economic


**Indicator Category**:

- [X] Syntheses of published information


**Synthesis Theme**:

- [X] Multiple System Drivers


```{r download_aquaculture}
# Download dataset from ecodata
try(downloadthis::download_this(
ecodata::aquaculture,
output_name = 'aquaculture',
output_extension = '.csv',
button_label = 'Download dataset from `ecodata`',
button_type = 'default',
has_icon = TRUE,
icon = 'fa fa-save',
class = 'hvr-sweep-to-left',
csv2 = F),
silent = TRUE)
```

## Introduction to Indicator
Aquaculture can produce seafood as well as other products. Here we summarize available data on aquaculture production from the Northeast US.

## Key Results and Visualizations
Overall, oyster production has generally increased and production per area has also improved. Aquaculture reporting is state specific, which can make it difficult to collate in a meaningful way. However, reporting coverage is improving year to year as aquaculture increases across the coast. The data included here only include oyster numbers at this time. Data after 2017 have decreased but, due to a lack of data from recent years, it is unclear if this has continued.

**Mid-Atlantic**
```{r plot_aquacultureMidAtlantic}
# Plot indicator
ggplotObject <- ecodata::plot_aquaculture(report = 'MidAtlantic', n = 10) +
ggplot2::ggtitle('')
ggplotObject
try(downloadthis::download_this(
ggplotObject$data,
output_name = 'aquacultureMidAtlantic',
output_extension = '.csv',
button_label = 'Download the data plotted above',
button_type = 'default',
has_icon = TRUE,
icon = 'fa fa-save',
class = 'hvr-sweep-to-left',
csv2 = F),
silent = TRUE)
```
<br>

**New England**
```{r plot_aquacultureNewEngland}
# Plot indicator
ggplotObject <- ecodata::plot_aquaculture(report = 'NewEngland', n = 10) +
ggplot2::ggtitle('')
ggplotObject
try(downloadthis::download_this(
ggplotObject$data,
output_name = 'aquacultureNewEngland',
output_extension = '.csv',
button_label = 'Download the data plotted above',
button_type = 'default',
has_icon = TRUE,
icon = 'fa fa-save',
class = 'hvr-sweep-to-left',
csv2 = F),
silent = TRUE)
```
<br>

## Implications
Aquaculture production contributes to overall seafood production in the Northeast US. This indicator provides only a portion of aquaculture production in the region, but increases in this portion are apparent across the whole time series. Recent decreases (since 2017) may indicate a downward change in this aquaculture portion; however, there is not enough data from years following 2017 to indicate any significant decreasing trends.

## Indicator statistics 
Spatial scale: Mid Atlantic and New England

Temporal scale: Annual

**Variable definitions**

```{r vars_aquaculture}
# Pull all var names
vars <- ecodata::aquaculture |>
   dplyr::select(Var) |>
   dplyr::distinct()

DT::datatable(vars)
```
## Get the data

**Point of contact**: Chris Schillaci <christopher.schillaci@noaa.gov>

**ecodata dataset**: `ecodata::aquaculture`

**Tech Doc link**: <https://noaa-edab.github.io/tech-doc/aquaculture.html>

### Public Availability

Source data are publicly available.

