# Aquaculture Production {#aquaculture}

**Description**: Oyster production: number of oysters harvested from aquaculture.

**Indicator family**: 

- [X] Megafauna
- [X] Social
- [X] Economic


**Contributor(s)**: Christopher Schillaci, Maine DMR, NH DES, MA DMF, RI CRMC, MD DNR

**Affiliations**: NOAA

```{r echo=FALSE}
knitr::opts_chunk$set(echo = F)
library(ecodata)
```
## Introduction to Indicator
Aquaculture can produce seafood as well as other products. Here we summarize available data on aquaculture production from Northeast US states.

## Key Results and Visualizations
Overall oyster production is increasing and production per area is also improving.  Aquaculture reporting is state specific and because of this can be difficult to collate in a meaningful way. However, this is improving year to year as aquaculture increases across the coast. The data included here only holds oyster numbers at this time.

```{r plot_aquacultureMAB}
# Plot indicator
ggplotObject <- ecodata::plot_aquaculture(report='MidAtlantic')
ggplotObject
```


## Indicator statistics 
Spatial scale: Mid Atlantic and New England

Temporal scale: Annual

**Synthesis Theme**:

- [X] Multiple System Drivers


```{r autostats_aquaculture}
# Either from Contributor or ecodata
```

## Implications
Aquaculture production contributes to overall seafood production in the Northeast US. This indicator provides only a portion of aquaculture production in the region, but increases in this portion are apparent.

## Get the data

**Point of contact**: [Chris Schillaci <christopher.schillaci@noaa.gov>](mailto:Chris Schillaci <christopher.schillaci@noaa.gov>){.email}

**ecodata name**: `ecodata::aquaculture`

**Variable definitions**

Pieces: number of oysters produced (all regions)                 
Shellfish lease Acres: area used for shellfish production (New England states only), acres  
Production/Acre: Pieces divided by Shellfish lease acres (New England states only)

```{r vars_aquaculture}
# Pull all var names
vars <- ecodata::aquaculture |>
   dplyr::select(Var) |>
   dplyr::distinct()

DT::datatable(vars)
```
**Indicator Category**:

- [X] Syntheses of published information


## Public Availability

Source data are publicly available.

## Accessibility and Constraints

_No response_

**tech-doc link**
<https://noaa-edab.github.io/tech-doc/aquaculture.html>

