# Bottom temperature - Seasonal Anomaly {#bottom_temp_comp}

**Description**: The data are seasonal bottom temperature anomaly time series for each EPU

**Indicator family**: 

- [X] Oceanographic


**Contributor(s)**: Joseph Caracappa, Hubert duPontavice, Vincent Saba, Zhuomin Chen

**Affiliations**: NEFSC

```{r echo=FALSE}
knitr::opts_chunk$set(echo = F)
library(ecodata)
```
## Introduction to Indicator
The raw bottom temperature product is in a horizontal 1/12 degree grid between 1959 and 2023 and is made of daily bottom temperature estimates from:

Bias-corrected ROMS-NWA (ROMScor) between 1959 and 1992 which was regridded in the same 1/12degree grid as GLORYS using bilinear interpolation; GLORYS12v1 in its original 1/12 degree grid between 1993 and 2023-08-29; and PSY in it's 1/12 degree grid from 2023-08-29 to 2023-12-31. 

Anomalies are calculated using the 1990-2020 reference period.

## Key Results and Visualizations
Time series plots for seasonal bottom temperature anomaly for each EPU shows a long-term warming trend. PSY forecasted data also included for comparison and indicates varying skill in predicting regional bottom temperature.

### MidAtlantic

```{r plot_bottom_temp_compMidAtlanticseasonalMAB}
# Plot indicator
ggplotObject <- ecodata::plot_bottom_temp_comp(report= 'MidAtlantic', varName= 'seasonal' ,EPU= 'MAB')
ggplotObject
```

```{r plot_bottom_temp_compMidAtlanticannualMAB}
# Plot indicator
ggplotObject <- ecodata::plot_bottom_temp_comp(report= 'MidAtlantic', varName= 'annual' ,EPU= 'MAB')
ggplotObject
```

### NewEngland

```{r plot_bottom_temp_compNewEnglandseasonalGB}
# Plot indicator
ggplotObject <- ecodata::plot_bottom_temp_comp(report= 'NewEngland', varName= 'seasonal' ,EPU= 'GB')
ggplotObject
```

```{r plot_bottom_temp_compNewEnglandseasonalGOM}
# Plot indicator
ggplotObject <- ecodata::plot_bottom_temp_comp(report= 'NewEngland', varName= 'seasonal' ,EPU= 'GOM')
ggplotObject
```

```{r plot_bottom_temp_compNewEnglandannualGB}
# Plot indicator
ggplotObject <- ecodata::plot_bottom_temp_comp(report= 'NewEngland', varName= 'annual' ,EPU= 'GB')
ggplotObject
```

```{r plot_bottom_temp_compNewEnglandannualGOM}
# Plot indicator
ggplotObject <- ecodata::plot_bottom_temp_comp(report= 'NewEngland', varName= 'annual' ,EPU= 'GOM')
ggplotObject
```


## Indicator statistics 
Spatial scale: EPU

Temporal scale: annual-seasonal

**Synthesis Theme**:

- [X] Multiple System Drivers


```{r autostats_bottom_temp_comp}
# Either from Contributor or ecodata
```

## Implications
Bottom temperature is an important driver for benthic and demersal species growth, metabolism, and reproduction. Changes in seasonal bottom temperature have implications for species' phenology

## Get the data

**Point of contact**: [joseph.caracappa@noaa.gov](mailto:joseph.caracappa@noaa.gov){.email}

**ecodata name**: `ecodata::bottom_temp_comp`

**Variable definitions**

Season: 1 = winter (January – March), 2 = spring (April – June), 3 = summer (July – September), 4 = fall (October – December) 
Subarea: EPU name 
Source:  ROMS (bias-corrected ROMS-NWA bottom temperature [@dupontavice_ocean_2022]), GLORYS (CMEM’s GLORYS12V1 global reanalysis bottom temperature), PSY (CMEM’s PSY global forecast bottom temperature) 
bt_temp : mean bottom temperature for each year/season across entire EPU 
ref_bt: bottom temperature climatology for season/EPU based on 1990-2020

```{r vars_bottom_temp_comp}
# Pull all var names
vars <- ecodata::bottom_temp_comp |>
   dplyr::select(Var, Units) |>
   dplyr::distinct()

DT::datatable(vars)
```
**Indicator Category**:

- [X] Syntheses of published information
- [X] Database pull with analysis


## Public Availability

Source data are publicly available.

## Accessibility and Constraints

_No response_

**tech-doc link**
<https://noaa-edab.github.io/tech-doc/bottom_temp_comp.html>

