# Bottom Temperature - in situ

**Description**: The data presented here are time series of regional average bottom temperature anomalies from ship-based measurements made on the Northeast Continental Shelf.

**Indicator family**: 

- [X] Oceanographic


**Contributor(s)**: Paula Fratantoni; Chris Melrose; Tamara Holzwarth-Davis

**Affiliations**: NEFSC

```{r echo=FALSE}
knitr::opts_chunk$set(echo = F)
library(ecodata)
```
## Introduction to Indicator
The bottom temperature index incorporates near-bottom temperature measurements collected on Northeast Fisheries Science Center (NEFSC) surveys between 1977-present. Early measurements were made using surface bucket samples, mechanical bathythermographs and expendable bathythermograph probes, but by 1991 the CTD – an acronym for conductivity temperature and depth – became standard equipment on all NEFSC surveys. Near-bottom refers to the deepest observation at each station that falls within 10 m of the reported water depth. Observations encompass the entire continental shelf area extending from Cape Hatteras, NC to Nova Scotia, Canada, inclusive of the Gulf of Maine and Georges Bank.

## Key Results and Visualizations
_No response_

### MAB

```{r plot_bottom_tempMAB}
# Plot indicator
ggplotObject <- ecodata::plot_bottom_temp(report='MidAtlantic')
print(ggplotObject)
```

### NE

```{r plot_bottom_tempNE}
# Plot indicator
ggplotObject <- ecodata::plot_bottom_temp(report='NewEngland')
print(ggplotObject)
```


## Indicator statistics 
Spatial scale: by EPU

Temporal scale: Annual

**Synthesis Theme**:




```{r autostats_bottom_temp}
# Either from Contributor or ecodata
```

## Implications
_No response_

## Get the data

**Point of contact**: [Paula.Fratantoni@noaa.gov](mailto:Paula.Fratantoni@noaa.gov){.email}

**ecodata name**: `ecodata::bottom_temp`

**Variable definitions**

Tbot_anom; Definition: Bottom temperature anomaly; Units: degree Celsius

```{r vars_bottom_temp}
# Pull all var names
vars <- ecodata::bottom_temp |>
   dplyr::select(Var, Units) |>
   dplyr::distinct()

DT::datatable(vars)
```
**Indicator Category**:

- [X] Extensive analysis, not yet published


## Public Availability

Source data are publicly available.

## Accessibility and Contraints

_No response_

**tech-doc link**
<https://noaa-edab.github.io/tech-doc/bottom_temp.html>

## References

