# Bottom temperature - High Resolution

**Description**: The data are gridded (1/12 degree) seasonal mean bottom temperature for the northeast U.S. shelf.

**Indicator family**: 

- [X] Oceanographic
- [ ] Habitat
- [ ] Lower trophic levels
- [ ] Megafauna
- [ ] Social
- [ ] Economic

**Contributor(s)**: Joseph Caracappa, Hubert duPontavice, Vincent Saba, Zhuomin Chen

**Affiliations**: NEFSC

```{r echo=FALSE}
knitr::opts_chunk$set(echo = F)
library(ecodata)
```
## What is this?
The bottom temperature product is in a horizontal 1/12 degree grid between 1959 and 2022 and is made of daily bottom temperature estimates from:

Bias-corrected ROMS-NWA (ROMScor) between 1959 and 1992 which was regridded in the same 1/12degree grid as GLORYS using bilinear interpolation; GLORYS12v1 in its original 1/12 degree grid between 1993 and 2020; GLO12v3 (called PSY4V3R1 in “A High-Resolution Ocean Bottom Temperature Product for the Northeast u.s. Continental Shelf Marine Ecosystem” ([2023](https://noaa-edab.github.io/tech-doc/bottom-temperature---high-resolution.html#ref-duPontavice2023)) and Lellouche et al. ([2018](https://noaa-edab.github.io/tech-doc/bottom-temperature---high-resolution.html#ref-Lellouche2018))) in its original 1/12 degree grid for 2021 to 2023.

### Visualizations
Proceed as though this were a short summary of a typical results section in a paper.

```{r plot_bottom_temp_seasonal_griddedMAB}
# Plot indicator
ggplotObject <- ecodata::plot_bottom_temp_seasonal_gridded(shadedRegion=c(2013,2022))
print(ggplotObject)
```
```{r plot_bottom_temp_seasonal_griddedNE}
# Plot indicator
ggplotObject <- ecodata::plot_bottom_temp_seasonal_gridded(shadedRegion=c(2013,2022),report='NewEngland')
print(ggplotObject)
```
### Indicator statistics 
Spatial scale: full shelf

Temporal scale: annual-seasonal

**Synthesis Theme**:

- [X] Multiple System Drivers
- [ ] Regime Shifts
- [ ] Ecosystem Reorganization

```{r autostats_bottom_temp_seasonal_gridded}
# Either from Contributor or ecodata
```

### Implications
_No response_

## Get the data

**Point of contact**: [joseph.caracappa@noaa.gov](joseph.caracappa@noaa.gov){.email}

**ecodata name**: `ecodata::bottom_temp_seasonal_gridded`

**variable names**

```{r vars_bottom_temp_seasonal_gridded}
# Pull all var names
vars <- ecodata::bottom_temp_seasonal_gridded |>
   dplyr::select(Var) |>
   dplyr::distinct()

DT::datatable(vars)
```
**Indicator Category**:

- [ ] Published Methods
- [ ] Extensive analysis, not yet published
- [X] Syntheses of published information
- [ ] Database pull
- [X] Database pull with analysis
- [ ] Other

### Public Availability

Source data are publicly available.

### Accessibility and Contraints

_No response_

**tech-doc link**
<https://noaa-edab.github.io/tech-doc/bottom_temp_seasonal_gridded.html>

## References

