# Bottom Temperature - in situ {#bottom_temp_insitu}

```{r echo=FALSE}
knitr::opts_chunk$set(echo = F)
library(ecodata)
library(downloadthis)
```

*Last updated July 23, 2026*

**Description**: The data presented here are time series of regional average bottom temperature anomalies from ship-based measurements made on the Northeast Continental Shelf.

**Contributor(s)**: Paula Fratantoni; Chris Melrose; Tamara Holzwarth-Davis

**Affiliations**: NEFSC

**Indicator Family**: 

- [X] Oceanographic


**Indicator Category**:

- [X] Extensive analysis, not yet published


**Synthesis Theme**:

- [X] Multiple System Drivers


```{r download_bottom_temp_insitu}
# Download dataset from ecodata
try(downloadthis::download_this(
ecodata::bottom_temp_insitu,
output_name = 'bottom_temp_insitu',
output_extension = '.csv',
button_label = 'Download dataset from `ecodata`',
button_type = 'default',
has_icon = TRUE,
icon = 'fa fa-save',
class = 'hvr-sweep-to-left',
csv2 = F),
silent = TRUE)
```

## Introduction to Indicator
The bottom temperature index incorporates near-bottom temperature measurements collected on Northeast Fisheries Science Center (NEFSC) surveys between 1977-present. Early measurements were made using surface bucket samples, mechanical bathythermographs and expendable bathythermograph probes, but by 1991 the CTD – an acronym for conductivity temperature and depth – became standard equipment on all NEFSC surveys. Near-bottom refers to the deepest observation at each station that falls within 10 m of the reported water depth. Observations encompass the entire continental shelf area extending from Cape Hatteras, NC to Nova Scotia, Canada, inclusive of the Gulf of Maine and Georges Bank.

## Key Results and Visualizations
_No response_

**Mid-Atlantic**
```{r plot_bottom_temp_insituMidAtlantic}
# Plot indicator
ggplotObject <- ecodata::plot_bottom_temp_insitu(report = 'MidAtlantic', n = 10) +
ggplot2::ggtitle('')
ggplotObject
try(downloadthis::download_this(
ggplotObject$data,
output_name = 'bottom_temp_insituMidAtlantic',
output_extension = '.csv',
button_label = 'Download the data plotted above',
button_type = 'default',
has_icon = TRUE,
icon = 'fa fa-save',
class = 'hvr-sweep-to-left',
csv2 = F),
silent = TRUE)
```
<br>

**New England**
```{r plot_bottom_temp_insituNewEngland}
# Plot indicator
ggplotObject <- ecodata::plot_bottom_temp_insitu(report = 'NewEngland', n = 10) +
ggplot2::ggtitle('')
ggplotObject
try(downloadthis::download_this(
ggplotObject$data,
output_name = 'bottom_temp_insituNewEngland',
output_extension = '.csv',
button_label = 'Download the data plotted above',
button_type = 'default',
has_icon = TRUE,
icon = 'fa fa-save',
class = 'hvr-sweep-to-left',
csv2 = F),
silent = TRUE)
```
<br>

## Implications
_No response_

## Indicator statistics 
Spatial scale: by EPU

Temporal scale: Annual

**Variable definitions**

```{r vars_bottom_temp_insitu}
# Pull all var names
vars <- ecodata::bottom_temp_insitu |>
   dplyr::select(Var, Units) |>
   dplyr::distinct()

DT::datatable(vars)
```
## Get the data

**Point of contact**: paula.fratantoni@noaa.gov

**ecodata dataset**: `ecodata::bottom_temp_insitu`

**Tech Doc link**: <https://noaa-edab.github.io/tech-doc/bottom_temp_insitu.html>

### Public Availability

Source data are publicly available.

