# Bottom temperature - Seasonal Anomaly

**Description**: The data are seasonal bottom temperature anomaly time series for each EPU

**Indicator family**: 

- [X] Oceanographic
- [ ] Habitat
- [ ] Lower trophic levels
- [ ] Megafauna
- [ ] Social
- [ ] Economic

**Contributor(s)**: Joseph Caracappa, Hubert duPontavice, Vincent Saba, Zhuomin Chen

**Affiliations**: NEFSC

```{r echo=FALSE}
knitr::opts_chunk$set(echo = F)
library(ecodata)
```
## Indroduction to Indicator
The raw bottom temperature product is in a horizontal 1/12 degree grid between 1959 and 2023 and is made of daily bottom temperature estimates from:

Bias-corrected ROMS-NWA (ROMScor) between 1959 and 1992 which was regridded in the same 1/12degree grid as GLORYS using bilinear interpolation; GLORYS12v1 in its original 1/12 degree grid between 1993 and 2023-08-29; and PSY in it's 1/12 degree grid from 2023-08-29 to 2023-12-31. 

Anomalies are calculated using the 1990-2020 reference period.

## Key Results and Visualizations
Time series plots for seasonal bottom temperature anomaly for each EPU shows a long-term warming trend. PSY forecasted data also included for comparison and indicates varying skill in predicting regional bottom temperature.

### MAB

```{r plot_bottom_temp_compMAB}
# Plot indicator
ggplotObject <- ecodata::plot_bottom_temp_comp(report='MidAtlantic')
print(ggplotObject)
```

### NE

```{r plot_bottom_temp_compNE}
# Plot indicator
ggplotObject <- ecodata::plot_bottom_temp_comp(report='NewEngland')
print(ggplotObject)
```


## Indicator statistics 
Spatial scale: EPU

Temporal scale: annual-seasonal

**Synthesis Theme**:

- [X] Multiple System Drivers
- [ ] Regime Shifts
- [ ] Ecosystem Reorganization

```{r autostats_bottom_temp_comp}
# Either from Contributor or ecodata
```

## Implications
Bottom temperature is an important driver for benthic and demersal species growth, metabolism, and reproduction. Changes in seasonal bottom temperature have implications for species' phenology

## Get the data

**Point of contact**: [joseph.caracappa@noaa.gov](joseph.caracappa@noaa.gov){.email}

**ecodata name**: `ecodata::bottom_temp_comp`

**Variable definitions**

Season: 1 = winter (January – March), 2 = spring (April – June), 3 = summer (July – September), 4 = fall (October – December) 
Subarea: EPU name 
Source:  ROMS (bias-corrected ROMS-NWA bottom temperature from du Pontavice et al, 2023), GLORYS (CMEM’s GLORYS12V1 global reanalysis bottom temperature), PSY (CMEM’s PSY global forecast bottom temperature) 
bt_temp : mean bottom temperature for each year/season across entire EPU 
ref_bt: bottom temperature climatology for season/EPU based on 1990-2020

```{r vars_bottom_temp_comp}
# Pull all var names
vars <- ecodata::bottom_temp_comp |>
   dplyr::select(Var, Units) |>
   dplyr::distinct()

DT::datatable(vars)
```
**Indicator Category**:

- [ ] Published Methods
- [ ] Extensive analysis, not yet published
- [X] Syntheses of published information
- [ ] Database pull
- [X] Database pull with analysis
- [ ] Other

## Public Availability

Source data are publicly available.

## Accessibility and Contraints

_No response_

**tech-doc link**
<https://noaa-edab.github.io/tech-doc/bottom_temp_comp.html>

## References

