# Calanus stage abundance anomaly {#calanus_stage}

```{r echo=FALSE}
knitr::opts_chunk$set(echo = F)
library(ecodata)
library(downloadthis)
```

*Last updated July 23, 2026*

**Description**: The data presented here are average annual abundance anomalies for different life stages of the copepod Calanus finmarcicus. Abundance anomaly data are presented for total, adult, and copepodite stages C1-C5. Also presented here are annual anomalies of surface and bottom salinity and temperature collected in situ.

**Contributor(s)**: Ryan Morse, Harvey Walsh

**Affiliations**: NEFSC

**Indicator Family**: 

- [x] Lower trophic levels


**Indicator Category**:

- [x] Published Methods
- [x] Extensive analysis, not yet published


**Synthesis Theme**:

- [x] Multiple System Drivers
- [x] Regime Shifts
- [x] Ecosystem Reorganization


```{r download_calanus_stage}
# Download dataset from ecodata
try(downloadthis::download_this(
ecodata::calanus_stage,
output_name = 'calanus_stage',
output_extension = '.csv',
button_label = 'Download dataset from `ecodata`',
button_type = 'default',
has_icon = TRUE,
icon = 'fa fa-save',
class = 'hvr-sweep-to-left',
csv2 = F),
silent = TRUE)
```

## Introduction to Indicator
Calanus finmarchicus is a dominant large-bodied copepod and serves as an important link between phytoplankton and higher trophic levels. In the Gulf of Maine, fluctuations in Calanus abundance have been linked to recruitment failures in commercial fish species (Perretti et al 2017), changes in fish condition (Smith et al. in prep), and distribution of right whales (Sorochan et al 2019). Here we present annual abundance anomalies of Calanus for copepodite life stages C1-C5, for mature adult copepods, and total abundance, along with in situ temperature and salinity at both surface and bottom.

## Key Results and Visualizations
Abundance anomalies for Calanus finmarchicus adult and stage C5 copepodite in the GOM were positive for 2023, following two years of negative anomalies

**Mid-Atlantic**
```{r plot_calanus_stageMidAtlantic}
# Plot indicator
ggplotObject <- ecodata::plot_calanus_stage(report = 'MidAtlantic') +
ggplot2::ggtitle('')
ggplotObject
try(downloadthis::download_this(
ggplotObject$data,
output_name = 'calanus_stageMidAtlantic',
output_extension = '.csv',
button_label = 'Download the data plotted above',
button_type = 'default',
has_icon = TRUE,
icon = 'fa fa-save',
class = 'hvr-sweep-to-left',
csv2 = F),
silent = TRUE)
```
<br>

**New England**
```{r plot_calanus_stageNewEngland}
# Plot indicator
ggplotObject <- ecodata::plot_calanus_stage(report = 'NewEngland') +
ggplot2::ggtitle('')
ggplotObject
try(downloadthis::download_this(
ggplotObject$data,
output_name = 'calanus_stageNewEngland',
output_extension = '.csv',
button_label = 'Download the data plotted above',
button_type = 'default',
has_icon = TRUE,
icon = 'fa fa-save',
class = 'hvr-sweep-to-left',
csv2 = F),
silent = TRUE)
```
<br>

## Implications
Despite the recent trend of high bottom temperature anomalies and low Calanus abundance, values for the most recent year (2023) show positive values for both adult and stage C5 copepodites, the largest copepodite stage before maturity. Higher Calanus abundance should be beneficial to right whales and other organisms that graze on Calanus.

## Indicator statistics 
Spatial scale: By EPU and full shelf

Temporal scale: Annual

**Variable definitions**

```{r vars_calanus_stage}
# Pull all var names
vars <- ecodata::calanus_stage |>
   dplyr::select(Var, Units) |>
   dplyr::distinct()

DT::datatable(vars)
```
## Get the data

**Point of contact**: Ryan Morse (ryan.morse@noaa.gov)

**ecodata dataset**: `ecodata::calanus_stage`

**Tech Doc link**: <https://noaa-edab.github.io/tech-doc/calanus_stage.html>

### Public Availability

Source data are NOT publicly available.

### Accessibility Constraints

Please email Harvey.Walsh@noaa.gov for further information and queries of copepod staged abundance data

