# Calanus Stage

**Description**: Calanus abundance per life stage 

**Contributor(s)**: Ryan Morse
  
**Data steward**: Ryan Morse <ryan.morse@noaa.gov>
  
**Point of contact**:  Ryan Morse <ryan.morse@noaa.gov>

## Impact Summary

Calanus finmarchicus is the most numerically abundant zooplankton taxa present in the waters of the GOM and GB during spring (Feb-Apr) (@Morse2017). The GOM is the southern limit of the viable reproduction in the range of this subarctic species, although it is often seasonally present in the MAB. Because of its thermal tolerance limitations as a subarctic species, it is projected that Calanus abundance will be drastically reduced in the GOM by the later half of this century if carbon emissions continue to increase as expected (@Grieve2017).

The life cycle of C. finmarchicus is complex, undergoing six nauplii stages and five copepodite stages before reaching maturity as an adult and reproducing (@Marshall1972). This complex lifecycle involves molting between stages and typically also involves a resting stage termed diapause, where immature copepodite stage-5 individuals migrate into deep basins and become dormant at depth during the late fall into winter. As a population present in waters across the North Atlantic, diapause is typically the rule and reproduction results in one generation per year (@Head2000), however, flexibility in maturation rates in response to environmental conditions and food availability allows Calanus to bypass the diapause phase and can result in several generations of Calanus within a year, particularly on GB (@Durbin1997). These local dynamics add to the variability of Calanus abundance in the NES system.

North Atlantic right whales (NARW) are a primary predator of Calanus finmarchicus. Due to a reduction in the filter feeding efficiency below ~ 1.5 mm (@Mayo2020), NARW primarily feed on adult and late stage copepodites (c4, c5). Typically only stage-5 copepodites undergo diapause and thus are often the bulk of prey available to NARW. In the GOM, diapusing Calanus c5 are typically found in the deep basins during late fall and winter. 

In the GOM during all seasons from 2000-2010, C. finmarchicus had high levels of abundance, coinciding with a period of strong recovery for the NARW population. That recovery plateaued after 2010 and has since been in decline. Adult C. finmarchicus levels reached time series maxima between 2000-2010 in the GOM, and the abundance of stage c5 copepodites also reached time series maxima in both spring and fall during this period. Following 2010, the abundance of Calanus plummeted in the GOM and a regime shift to lower biomass was detected for both GOM and GBK (Sorochan et al. 2019). NARW abandoned their traditional feeding grounds in the GOM and were increasingly present in Canadian waters in the Gulf of St. Lawrence during summer and fall beginning in 2010 and peaking in 2015 (@Simard2019). This shift in distribution may be related to food availability and the abundance of larger species of Calanus in the GSL.


### Plotted data 
```{r, code = readLines("https://raw.githubusercontent.com/NOAA-EDAB/ecodata/master/chunk-scripts/LTL_NE.Rmd-gom-calanus-stage.R")}

```


### Raw data 
```{r}
library(DT)
datatable(ecodata::CalanusStage, options = list(pageLength = 10))
```

## Methods
### Data Sources

### Data Extraction

### Data Analysis
Code used to analyze calanus stage data can be found [at this link](https://github.com/NOAA-EDAB/tech-doc/blob/master/R/stored_scripts/CalanusStage_SOE.R). 

### Data Processing

Code for processing Calanus stage data can be found  [here](https://raw.githubusercontent.com/NOAA-EDAB/ecodata/master/chunk-scripts/LTL_MAB.Rmd-calanus-stage.R). 
