# Calanus Stage

**Description**: Calanus abundance per life stage 

**Contributor(s)**: Ryan Morse
  
**Data steward**: Ryan Morse <ryan.morse@noaa.gov>
  
**Point of contact**:  Ryan Morse <ryan.morse@noaa.gov>

## Impact Summary

Description of why this indicator is important and what is going on in the Northeast related to this information. 

### Plotted data 
```{r, code = readLines("https://raw.githubusercontent.com/NOAA-EDAB/ecodata/master/chunk-scripts/LTL_NE.Rmd-gb-calanus-stage.R")}

```


### Raw data 
```{r}
library(DT)
datatable(ecodata::CalanusStage, options = list(pageLength = 10))
```

## Methods
### Data Sources

### Data Extraction

### Data Analysis
Code used to analyze calanus stage data can be found [at this link](https://github.com/NOAA-EDAB/tech-doc/blob/master/R/stored_scripts/CalanusStage_SOE.R). 

### Data Processing

Code for processing Calanus stage data can be found  [here](https://raw.githubusercontent.com/NOAA-EDAB/ecodata/master/chunk-scripts/LTL_MAB.Rmd-calanus-stage.R). 
