# Chesapeake Bay Salinity {#ch_bay_sal}

**Description**: This data is collected from the CBIBS buoy system.

**Indicator family**: 

- [X] Oceanographic


**Contributor(s)**: Charles Pellerin

**Affiliations**: NEFSC

```{r echo=FALSE}
knitr::opts_chunk$set(echo = F)
library(ecodata)
```
## Introduction to Indicator
The indicator is the Water Temperature and the Water Salinity.

## Key Results and Visualizations
The key results are the state of the water temperature and the salinity

### MAB

```{r plot_ch_bay_salMAB}
# Plot indicator
ggplotObject <- ecodata::plot_ch_bay_sal(report='MidAtlantic')
ggplotObject
```


## Indicator statistics 
Spatial scale: Main stem of the Chesapeake Bay

Temporal scale: Annual

**Synthesis Theme**:

- [X] Multiple System Drivers


```{r autostats_ch_bay_sal}
# Either from Contributor or ecodata
```

## Implications
The changes in the temperature and salinity have implications in the habitat

## Get the data

**Point of contact**: [Charles Pellerin (charles.pellerin@noaa.gov)](mailto:Charles Pellerin (charles.pellerin@noaa.gov)){.email}

**ecodata name**: `ecodata::ch_bay_sal`

**Variable definitions**

1) Salinity

```{r vars_ch_bay_sal}
# Pull all var names
vars <- ecodata::ch_bay_sal |>
   dplyr::select(Var, Units) |>
   dplyr::distinct()

DT::datatable(vars)
```
**Indicator Category**:

- [X] Database pull


## Public Availability

Source data are publicly available.

## Accessibility and Constraints

_No response_

**tech-doc link**
<https://noaa-edab.github.io/tech-doc/ch_bay_sal.html>

