# Chesapeake Bay Water Quality Standards Attainment

**Description**: Chesapeake Bay Water Quality Attainment Indicator

**Indicator family**: 

- [X] Lower trophic levels


**Contributor(s)**: Qian Zhang, Richard Tian, and Peter Tango

**Affiliations**: ?

```{r echo=FALSE}
knitr::opts_chunk$set(echo = F)
library(ecodata)
```
## Introduction to Indicator
To protect the aquatic living resources of Chesapeake Bay, the Chesapeake Bay Program (CBP) partnership has developed a guidance framework of ambient water quality criteria with designated uses and assessment procedures for dissolved oxygen, chlorophyll-a, and water clarity/submerged aquatic vegetation (SAV) (U.S. Environmental Protection Agency, 2003). To achieve consistent assessment over time and between jurisdictions, a multimetric indicator was proposed by the CBP partnership to provide a means for tracking the progress in all 92 management segments of Chesapeake Bay (Hernandez-Cordero et al., 2020; U.S. Environmental Protection Agency, 2017). This indicator has been computed for each three-year assessment period since 1985-1987, providing an integrated measure of Chesapeake Bay’s water quality condition over the last three decades.

The multimetric indicator required monitoring data on dissolved oxygen (DO) concentrations, chlorophyll-a concentrations, water clarity, SAV acreage, water temperature, and salinity. SAV acreage has been measured by the Virginia Institute of Marine Science in collaboration with the CBP, which is available via http://web.vims.edu/bio/sav/StateSegmentAreaTable.htm. Data for all other parameters were obtained from the CBP Water Quality Database (http://www.chesapeakebay.net/data/downloads/cbp_water_quality_database_1984_present). These data have been routinely reported to the CBP by the Maryland Department of Natural Resources, Virginia Department of Environmental Quality, Old Dominion University, Virginia Institute of Marine Science, and citizen/volunteer monitoring initiatives.

Monitoring data of DO, chlorophyll-a, and water clarity/SAV were processed and compared with water quality criteria thresholds according to different designated uses (DUs). These DUs are migratory spawning and nursery (MSN), open water (OW), deep water (DW), deep channel (DC), and shallow water (SW), which reflect the seasonal nature of water column structure and the life history needs of living resources. Station-level DO and chlorophyll-a data were spatially interpolated in three dimensions. Salinity and water temperature data were used to compute the vertical density structure of the water column, which was translated into layers of different DUs. Criteria attainment was determined by comparing violation rates over a 3-year period to a reference cumulative frequency distribution that represents the extent of allowable violation. This approach was implemented using FORTRAN codes, which are provided as a zipped folder. For water clarity/SAV, the single best year in the 3-year assessment period was compared with the segment-specific acreage goal, the water clarity goal, or a combination of both. For more details, refer to the Methods section of Zhang et al. (2018).

The multimetric indicator quantifies the fraction of segment-DU-criterion combinations that meet all applicable season-specific thresholds for each 3-year assessment period from 1985-1987 to 2019-2021. For each 3-year assessment period, all applicable segment-DU-criterion combinations were evaluated in a binomial fashion and scored 1 for “in attainment” and 0 for “nonattainment”. The classified status of each segment-DU-criterion combination was weighted via segments’ surface area and summed to obtain the multimetric index score. This weighting scheme was adopted for two reasons: (1) segments vary in size over four orders of magnitude, and (2) surface area of each segment does not change with time or DUs, unlike seasonally variable habitat volume or bottom water area (U.S. Environmental Protection Agency, 2017). For more details, refer to the Methods section of Zhang et al. (2018).

Hernandez-Cordero AL, Tango PJ, Batiuk RA. Development of a multimetric water quality Indicator for tracking progress towards the achievement of Chesapeake Bay water quality standards. Environmental Monitoring and Assessment 2020; 192.

U.S. Environmental Protection Agency. Ambient water quality criteria for dissolved oxygen, water clarity and chlorophyll-a for the Chesapeake Bay and its tidal tributaries. USEPA Region III Chesapeake Bay Program Office, Annapolis, Maryland, 2003.

U.S. Environmental Protection Agency. Ambient water quality criteria for dissolved oxygen, water clarity and chlorophyll-a for the Chesapeake Bay and its tidal tributaries: 2017 addendum. USEPA Region III Chesapeake Bay Program Office, Annapolis, Maryland, 2017.

Zhang Q, Murphy RR, Tian R, Forsyth MK, Trentacoste EM, Keisman J, et al. Chesapeake Bay's water quality condition has been recovering: Insights from a multimetric indicator assessment of thirty years of tidal monitoring data. Science of the Total Environment 2018; 637-638: 1617-1625.

## Key Results and Visualizations
The indicator provides an integrated measure of Chesapeake Bay’s water quality condition (Figure 1). In 2019-2021, 28.1% of all tidal water segment-DU-criterion combinations are estimated to have met or exceeded applicable water quality criteria thresholds. Overall, the indicator has a positive and statistically significant trend between 1985-1987 and 2019-2021, which shows that Chesapeake Bay is on a positive trajectory toward recovery. This pattern has been statistically linked to total nitrogen reduction, indicating responsiveness of attainment status to management actions implemented to reduce nutrients (Zhang et al., 2018).

### MAB

```{r plot_ches_bay_wqMAB}
# Plot indicator
ggplotObject <- ecodata::plot_ches_bay_wq(report='MidAtlantic')
print(ggplotObject)
```


## Indicator statistics 
Spatial scale: Chesapeake Bay

Temporal scale: 3-year assessment period between 1985-1987 and 2019-2021.

**Synthesis Theme**:

- [X] Multiple System Drivers


```{r autostats_ches_bay_wq}
# Either from Contributor or ecodata
```

## Implications
Patterns of attainment of individual designated uses are variable (Figure 2). According to Mann-Kendall trend analysis extended from Zhang et al. (2018), dissolved oxygen criterion attainment (migratory fish spawning & nursery) showed a statistically significant long-term decline. By contrast, dissolved oxygen criterion attainment (specifically deep water and deep channel) and water clarity/SAV criterion attainment showed statistically significant long-term improvements, which may be an indication of increasing resilience in the bay ecosystem since the Bay restoration began.

## Get the data

**Point of contact**: [Qian Zhang, qzhang@chesapeakebay.net](mailto:Qian Zhang, qzhang@chesapeakebay.net){.email}

**ecodata name**: `ecodata::ches_bay_wq`

**Variable definitions**

Period: Assessment period Year 1: Starting year of the assessment period Year 2: Ending year of the assessment period 
Total: The overall attainment indicator MSN-DO: Estimated attainment of the dissolved oxygen criterion for the migratory spawning and nursery designated use 
OW-DO: Estimated attainment of the dissolved oxygen criterion for the open water designated use 
DW-DO: Estimated attainment of the dissolved oxygen criterion for the deep water designated use 
DC-DO: Estimated attainment of the dissolved oxygen criterion for the deep channel designated use OW-CHLA: Estimated attainment of the chlorophyll-a criterion 
SW-Clarity/SAV: Estimated attainment of the bay grasses / water clarity criterion for the shallow water designated use

```{r vars_ches_bay_wq}
# Pull all var names
vars <- ecodata::ches_bay_wq |>
   dplyr::select(Var, Units) |>
   dplyr::distinct()

DT::datatable(vars)
```
**Indicator Category**:

- [X] Database pull with analysis


## Public Availability

Source data are publicly available.

## Accessibility and Contraints

_No response_

**tech-doc link**
<https://noaa-edab.github.io/tech-doc/ches_bay_wq.html>

## References

