# Phytoplankton chlorophyll, size class and primary production

**Description**: Satellite derived phytoplankton data including chlorophyll concentration, phytoplankton size class, and primary production for the Northeast Continental Shelf and ecological production units.

(To be expanded)

**Indicator family**: 

- [ ] Oceanographic
- [ ] Habitat
- [X] Lower trophic levels
- [ ] Megafauna
- [ ] Social
- [ ] Economic

**Contributor(s)**: Kimberly Hyde

**Affiliations**: NEFSC

```{r echo=FALSE}
knitr::opts_chunk$set(echo = F)
library(ecodata)
```
## What is this?
Phytoplankton are the foundation of the marine food web and are the primary food source for zooplankton and filter feeders such as shellfish. Numerous environmental and oceanographic factors interact to drive the abundance, composition, spatial distribution, and productivity of phytoplankton. Satellite derived measurements of chlorophyll, the dominant photosynthetic pigment in phytoplankton, are used to estimate phytoplankton biomass. Phytoplankton growth depends on the availability of carbon dioxide, sunlight and nutrients and their growth rates can be influenced by water temperature, water depth, wind, and grazing pressure.  Primary productivity is a measure of the amount of carbon produced by phytoplankton. The seasonal cycle of phytoplankton size distribution are typically dominated by larger-celled microplankton during the winter-spring and fall bloom periods, while smaller-celled nanoplankton dominate during the warmer summer months.

(To be expanded)

### Visualizations
(In development)

```{r plot_chl_ppMAB}
# Plot indicator
ggplotObject <- ecodata::plot_chl_pp(shadedRegion=c(2013,2022))
print(ggplotObject)
```
```{r plot_chl_ppNE}
# Plot indicator
ggplotObject <- ecodata::plot_chl_pp(shadedRegion=c(2013,2022),report='NewEngland')
print(ggplotObject)
```
### Indicator statistics 
Spatial scale: By EPU and gridded for the entire shelf

Temporal scale: Daily, weekly, monthly, annual, climatology (1998 to current year)

**Synthesis Theme**:

- [X] Multiple System Drivers
- [ ] Regime Shifts
- [ ] Ecosystem Reorganization

```{r autostats_chl_pp}
# Either from Contributor or ecodata
```

### Implications
(In development)

## Get the data

**Point of contact**: [kimberly.hyde@noaa.gov](kimberly.hyde@noaa.gov){.email}

**ecodata name**: `ecodata::chl_pp`

**variable names**

```{r vars_chl_pp}
# Pull all var names
vars <- ecodata::chl_pp |>
   dplyr::select(Var, Units) |>
   dplyr::distinct()

DT::datatable(vars)
```
**Indicator Category**:

- [X] Published Methods
- [ ] Extensive analysis, not yet published
- [ ] Syntheses of published information
- [ ] Database pull
- [ ] Database pull with analysis
- [X] Other

**Indicator Category**:

Publicly available satellite data that are processed and analyzed

### Public Availability

Source data are publicly available.

### Accessibility and Contraints

_No response_

**tech-doc link**
<https://noaa-edab.github.io/tech-doc/chl_pp.html>

## References

