# Cold Pool Index

**Description**: Three annual cold pool indices (and standard error) for ss1959 through 2023

**Indicator family**: 

- [X] Oceanographic


**Contributor(s)**: Joe Caracappa, Hubert du Pontavice, Vincent Saba, Zhuomin Chen

**Affiliations**: NEFSC

```{r echo=FALSE}
knitr::opts_chunk$set(echo = F)
library(ecodata)
```
## Introduction to Indicator
The cold pool is seasonal feature in the Mid-Atlantic Bight that is defined by it's low temperature (< 10 deg C), relative freshness ( < 34 psu), and moderate depth (20 -200m). It is typically present between June and September. Cold pool dynamics can influence the recruitment and settlement of fish species. This indicator set shows the intensity, extent, and persistence of the cold pool each year based on gridded model and reanalysis products.

## Key Results and Visualizations
Time series plots of the three cold pool indices. Cold pool index shows the mean temperature within the cold pool where positive values indicate a warming cold pool. Cold pool extent shows the change in maximum area relative to the historical mean, where negative values indicate a shrinking cold pool. Cold pool persistence measures the duration of the cold pool relative to the historical mean. Negative values indicate a shorter duration. In general the cold pool has been getting warmer, has persisted for a shorter duration, and has covered a smaller footprint since the 1960s.

### MidAtlantic

```{r plot_cold_poolMidAtlanticcold_pool}
# Plot indicator
ggplotObject <- ecodata::plot_cold_pool(report= 'MidAtlantic', varName= 'cold_pool')
print(ggplotObject)
```

```{r plot_cold_poolMidAtlanticpersistence}
# Plot indicator
ggplotObject <- ecodata::plot_cold_pool(report= 'MidAtlantic', varName= 'persistence')
print(ggplotObject)
```

```{r plot_cold_poolMidAtlanticextent}
# Plot indicator
ggplotObject <- ecodata::plot_cold_pool(report= 'MidAtlantic', varName= 'extent')
print(ggplotObject)
```

### NewEngland

```{r plot_cold_poolNewEnglandcold_pool}
# Plot indicator
ggplotObject <- ecodata::plot_cold_pool(report= 'NewEngland', varName= 'cold_pool')
print(ggplotObject)
```

```{r plot_cold_poolNewEnglandpersistence}
# Plot indicator
ggplotObject <- ecodata::plot_cold_pool(report= 'NewEngland', varName= 'persistence')
print(ggplotObject)
```

```{r plot_cold_poolNewEnglandextent}
# Plot indicator
ggplotObject <- ecodata::plot_cold_pool(report= 'NewEngland', varName= 'extent')
print(ggplotObject)
```


## Indicator statistics 
Spatial scale: MAB

Temporal scale: annual

**Synthesis Theme**:

- [X] Multiple System Drivers


```{r autostats_cold_pool}
# Either from Contributor or ecodata
```

## Implications
Changes in cold pool indicators can be signs of changes in regional/seasonal oceanographic patterns. This may impact the recruitment and behavior of species dependent on the cold pool.

## Get the data

**Point of contact**: [joseph.caracappa@noaa.gov](mailto:joseph.caracappa@noaa.gov){.email}

**ecodata name**: `ecodata::cold_pool`

**Variable definitions**

1) Source: ROMS (bias-corrected ROMS-NWA bottom temperature from du Pontavice et al, 2023), GLORYS (CMEM’s GLORYS12V1 global reanalysis bottom temperature), PSY (CMEM’s PSY global forecast bottom temperature) 
2) year 3) cold_pool_index: measure of mean temperature within cold pool 
4) se_cold_pool_index: standard error of cold_pool_index 
5) persistence_index: measure of duration of cold pool 
6) se_persistence_index: standard error of persistence_index 
7) extent_index: measure of spatial extent of cold pool 
8) se_extent_index: standard error of extent_index

```{r vars_cold_pool}
# Pull all var names
vars <- ecodata::cold_pool |>
   dplyr::select(Var) |>
   dplyr::distinct()

DT::datatable(vars)
```
**Indicator Category**:

- [X] Published Methods
- [X] Syntheses of published information


## Public Availability

Source data are publicly available.

## Accessibility and Constraints

_No response_

**tech-doc link**
<https://noaa-edab.github.io/tech-doc/cold_pool.html>

## References

