# Commercial Catch and Fleet Diversity {#commercial_div}

```{r echo=FALSE}
knitr::opts_chunk$set(echo = F)
library(ecodata)
library(downloadthis)
```

*Last updated July 23, 2026*

**Description**: Permit-level species diversity and Council-level fleet diversity.

**Contributor(s)**: Geret DePiper, Min-Yang Lee

**Affiliations**: NEFSC

**Indicator Family**: 

- [X] Social
- [X] Economic


**Indicator Category**:

- [X] Published Methods


**Synthesis Theme**:

- [X] Multiple System Drivers
- [X] Regime Shifts
- [X] Ecosystem Reorganization


```{r download_commercial_div}
# Download dataset from ecodata
try(downloadthis::download_this(
ecodata::commercial_div,
output_name = 'commercial_div',
output_extension = '.csv',
button_label = 'Download dataset from `ecodata`',
button_type = 'default',
has_icon = TRUE,
icon = 'fa fa-save',
class = 'hvr-sweep-to-left',
csv2 = F),
silent = TRUE)
```

## Introduction to Indicator
Diversity estimates have been developed to understand whether specialization, or alternatively stovepiping, is occurring in fisheries of the Northeastern Large Marine Ecosystem. We use the average effective Shannon indices for species revenue at the permit level, for all permits landing any amount of [NEFMC](https://www.nefmc.org/) or [MAFMC](http://www.mafmc.org/) Fishery Management Plan (FMP) species within a year (including both Monkfish and Spiny Dogfish). We also use the effective Shannon index of fleet revenue diversity and count of active fleets to assess the extent to which the distribution of fishing changes across fleet segments.

## Key Results and Visualizations
Diversity estimates have been developed for species landed by commercial vessels with New England permits, and fleets landing managed species. Over the course of the last three years, there has been a steep decline in the effective number of species being landed in the commercial fleet, and current diversity is near a low since records began.  Commercial fishery fleet count has rebounded in recent years, although still at levels well below the historical average. Here a fleet is defined as the combination of gear type (Scallop Dredge, Clam Dredge, Other Dredge, Gillnet, Hand Gear, Longline, Bottom Trawl, Midwater Trawl, Pot, or Purse Seine) and vessel length category (Less than 30 ft, 30 to 50 ft, 50 to 75 feet, 75 ft and above). 

Commercial fishery fleet count and fleet diversity have been stable over time in the MAB, with current values near the long-term average. This indicates similar commercial fleet composition and species targeting opportunities over time. Commercial fisheries are relying on fewer species relative to the mid-90s, although current species revenue diversity has recovered somewhat in the last year.

**Fleet count, Mid-Atlantic**
```{r plot_commercial_divFleet countMidAtlantic}
# Plot indicator
ggplotObject <- ecodata::plot_commercial_div(varName = 'Fleet count', report = 'MidAtlantic', n = 10) +
ggplot2::ggtitle('')
ggplotObject
try(downloadthis::download_this(
ggplotObject$data,
output_name = 'commercial_divFleet countMidAtlantic',
output_extension = '.csv',
button_label = 'Download the data plotted above',
button_type = 'default',
has_icon = TRUE,
icon = 'fa fa-save',
class = 'hvr-sweep-to-left',
csv2 = F),
silent = TRUE)
```
<br>

**Fleet diversity in revenue, Mid-Atlantic**
```{r plot_commercial_divFleet diversity in revenueMidAtlantic}
# Plot indicator
ggplotObject <- ecodata::plot_commercial_div(varName = 'Fleet diversity in revenue', report = 'MidAtlantic', n = 10) +
ggplot2::ggtitle('')
ggplotObject
try(downloadthis::download_this(
ggplotObject$data,
output_name = 'commercial_divFleet diversity in revenueMidAtlantic',
output_extension = '.csv',
button_label = 'Download the data plotted above',
button_type = 'default',
has_icon = TRUE,
icon = 'fa fa-save',
class = 'hvr-sweep-to-left',
csv2 = F),
silent = TRUE)
```
<br>

**Permit revenue species diversity, Mid-Atlantic**
```{r plot_commercial_divPermit revenue species diversityMidAtlantic}
# Plot indicator
ggplotObject <- ecodata::plot_commercial_div(varName = 'Permit revenue species diversity', report = 'MidAtlantic', n = 10) +
ggplot2::ggtitle('')
ggplotObject
try(downloadthis::download_this(
ggplotObject$data,
output_name = 'commercial_divPermit revenue species diversityMidAtlantic',
output_extension = '.csv',
button_label = 'Download the data plotted above',
button_type = 'default',
has_icon = TRUE,
icon = 'fa fa-save',
class = 'hvr-sweep-to-left',
csv2 = F),
silent = TRUE)
```
<br>

**Fleet count, New England**
```{r plot_commercial_divFleet countNewEngland}
# Plot indicator
ggplotObject <- ecodata::plot_commercial_div(varName = 'Fleet count', report = 'NewEngland', n = 10) +
ggplot2::ggtitle('')
ggplotObject
try(downloadthis::download_this(
ggplotObject$data,
output_name = 'commercial_divFleet countNewEngland',
output_extension = '.csv',
button_label = 'Download the data plotted above',
button_type = 'default',
has_icon = TRUE,
icon = 'fa fa-save',
class = 'hvr-sweep-to-left',
csv2 = F),
silent = TRUE)
```
<br>

**Fleet diversity in revenue, New England**
```{r plot_commercial_divFleet diversity in revenueNewEngland}
# Plot indicator
ggplotObject <- ecodata::plot_commercial_div(varName = 'Fleet diversity in revenue', report = 'NewEngland', n = 10) +
ggplot2::ggtitle('')
ggplotObject
try(downloadthis::download_this(
ggplotObject$data,
output_name = 'commercial_divFleet diversity in revenueNewEngland',
output_extension = '.csv',
button_label = 'Download the data plotted above',
button_type = 'default',
has_icon = TRUE,
icon = 'fa fa-save',
class = 'hvr-sweep-to-left',
csv2 = F),
silent = TRUE)
```
<br>

**Permit revenue species diversity, New England**
```{r plot_commercial_divPermit revenue species diversityNewEngland}
# Plot indicator
ggplotObject <- ecodata::plot_commercial_div(varName = 'Permit revenue species diversity', report = 'NewEngland', n = 10) +
ggplot2::ggtitle('')
ggplotObject
try(downloadthis::download_this(
ggplotObject$data,
output_name = 'commercial_divPermit revenue species diversityNewEngland',
output_extension = '.csv',
button_label = 'Download the data plotted above',
button_type = 'default',
has_icon = TRUE,
icon = 'fa fa-save',
class = 'hvr-sweep-to-left',
csv2 = F),
silent = TRUE)
```
<br>

## Implications
Fleet diversity indices can be used to evaluate stability objectives as well as risks to fishery resilience and to maintaining equity in access to fishery resources [@gaichas_implementing_2018]. In New England, the relatively low diversity estimates for the commercial fishery are likely driven by the continued reliance on a few species, sea scallops and lobster. This trend could diminish the capacity to respond to future fishing opportunities. 

In the Mid-Atlantic, stability in commercial fleet diversity metrics suggests stable capacity to respond to the current range of fishing opportunities.

## Indicator statistics 
Spatial scale: Mid-Atlantic and New England

Temporal scale: Annual

**Variable definitions**

```{r vars_commercial_div}
# Pull all var names
vars <- ecodata::commercial_div |>
   dplyr::select(Var, Units) |>
   dplyr::distinct()

DT::datatable(vars)
```
## Get the data

**Point of contact**: EDAB, edab.data@noaa.gov

**ecodata dataset**: `ecodata::commercial_div`

**Tech Doc link**: <https://noaa-edab.github.io/tech-doc/commercial_div.html>

### Public Availability

Source data are NOT publicly available.

### Accessibility Constraints

Request data from GARFO

