# Dissolved Oxygen {#dissolved_oxygen}

**Description**: Coming soon

**Indicator family**: 

- [x] Oceanographic
- [x] Habitat


**Contributor(s)**: Coming soon

**Affiliations**: Coming soon

```{r echo=FALSE}
knitr::opts_chunk$set(echo = F)
library(ecodata)
```
## Introduction to Indicator
Coming soon

## Key Results and Visualizations
Coming soon

```{r plot_dissolved_oxygen}
# Plot indicator
ggplotObject <- ecodata::plot_dissolved_oxygen()
ggplotObject
```


## Indicator statistics 
Spatial scale: Coming soon

Temporal scale: Coming soon

**Synthesis Theme**:

- [x] Multiple System Drivers


```{r autostats_dissolved_oxygen}
# Either from Contributor or ecodata
```

## Implications
Coming soon

## Get the data

**Point of contact**: [Coming soon](mailto:Coming soon){.email}

**ecodata name**: `ecodata::dissolved_oxygen`

**Variable definitions**

Coming soon

```{r vars_dissolved_oxygen}
# Pull all var names
vars <- ecodata::dissolved_oxygen |>
   dplyr::select(Var) |>
   dplyr::distinct()

DT::datatable(vars)
```
**Indicator Category**:




## Public Availability

Source data are publicly available.

## Accessibility and Constraints

_No response_

**tech-doc link**
<https://noaa-edab.github.io/tech-doc/dissolved_oxygen.html>

