# NE Shelf Functional Trait Indicators {#finfish_traits}

```{r echo=FALSE}
knitr::opts_chunk$set(echo = F)
library(ecodata)
library(downloadthis)
```

*Last updated July 23, 2026*

**Description**: These data represent estimates of trait distributions for key functional traits amongst finfish species at each EPU on the Northeast Continental Shelf.

**Contributor(s)**: Bart DiFiore, Kathy Mills, Scott Large

**Affiliations**: Massachusetts

**Indicator Family**: 

- [X] Megafauna


**Indicator Category**:

- [X] Extensive analysis, not yet published
- [X] Database pull with analysis


**Synthesis Theme**:

- [X] Regime Shifts
- [X] Ecosystem Reorganization


```{r download_finfish_traits}
# Download dataset from ecodata
try(downloadthis::download_this(
ecodata::finfish_traits,
output_name = 'finfish_traits',
output_extension = '.csv',
button_label = 'Download dataset from `ecodata`',
button_type = 'default',
has_icon = TRUE,
icon = 'fa fa-save',
class = 'hvr-sweep-to-left',
csv2 = F),
silent = TRUE)
```

## Introduction to Indicator
Functional traits, such as length at maturity, asymptotic body size, or fecundity, offer a means to synthesize change across complex, diverse communities while avoiding the pitfalls of examining change in any given species. In general, ecological theory predicts that if many species play similar roles in an ecosystem, known as high functional redundancy, then the loss of any one species will be compensated by the prevalence of functionally similar species, allowing for greater stability and resilience at the ecosystem level. Therefore, monitoring changes in functional trait distributions can provide a means of assessing ecosystem-scale resilience. 

Here, we compiled eight key functional traits and three aggregate trait ordinations for ~388 finfish species encountered in the NEFSC trawl survey. We then estimated the prevalence of each trait for each season, year, and EPU, using a community-weighed mean approach. Finally, we used regression-based approaches to explore long-term temporal shifts in functional trait distributions.

## Key Results and Visualizations
[SOE_Submission_results.docx](https://github.com/user-attachments/files/17590569/SOE_Submission_results.docx)

I'm happy to share the code used to produce these figures from the data linked above.

**Mid-Atlantic, trophic_level**
```{r plot_finfish_traitsMidAtlantictrophic_level}
# Plot indicator
ggplotObject <- ecodata::plot_finfish_traits(report = 'MidAtlantic', varName = 'trophic_level', n = 10) +
ggplot2::ggtitle('')
ggplotObject
try(downloadthis::download_this(
ggplotObject$data,
output_name = 'finfish_traitsMidAtlantictrophic_level',
output_extension = '.csv',
button_label = 'Download the data plotted above',
button_type = 'default',
has_icon = TRUE,
icon = 'fa fa-save',
class = 'hvr-sweep-to-left',
csv2 = F),
silent = TRUE)
```
<br>

**New England, trophic_level**
```{r plot_finfish_traitsNewEnglandtrophic_level}
# Plot indicator
ggplotObject <- ecodata::plot_finfish_traits(report = 'NewEngland', varName = 'trophic_level', n = 10) +
ggplot2::ggtitle('')
ggplotObject
try(downloadthis::download_this(
ggplotObject$data,
output_name = 'finfish_traitsNewEnglandtrophic_level',
output_extension = '.csv',
button_label = 'Download the data plotted above',
button_type = 'default',
has_icon = TRUE,
icon = 'fa fa-save',
class = 'hvr-sweep-to-left',
csv2 = F),
silent = TRUE)
```
<br>

**Mid-Atlantic, offspring_size**
```{r plot_finfish_traitsMidAtlanticoffspring_size}
# Plot indicator
ggplotObject <- ecodata::plot_finfish_traits(report = 'MidAtlantic', varName = 'offspring_size', n = 10) +
ggplot2::ggtitle('')
ggplotObject
try(downloadthis::download_this(
ggplotObject$data,
output_name = 'finfish_traitsMidAtlanticoffspring_size',
output_extension = '.csv',
button_label = 'Download the data plotted above',
button_type = 'default',
has_icon = TRUE,
icon = 'fa fa-save',
class = 'hvr-sweep-to-left',
csv2 = F),
silent = TRUE)
```
<br>

**New England, offspring_size**
```{r plot_finfish_traitsNewEnglandoffspring_size}
# Plot indicator
ggplotObject <- ecodata::plot_finfish_traits(report = 'NewEngland', varName = 'offspring_size', n = 10) +
ggplot2::ggtitle('')
ggplotObject
try(downloadthis::download_this(
ggplotObject$data,
output_name = 'finfish_traitsNewEnglandoffspring_size',
output_extension = '.csv',
button_label = 'Download the data plotted above',
button_type = 'default',
has_icon = TRUE,
icon = 'fa fa-save',
class = 'hvr-sweep-to-left',
csv2 = F),
silent = TRUE)
```
<br>

**Mid-Atlantic, age_maturity**
```{r plot_finfish_traitsMidAtlanticage_maturity}
# Plot indicator
ggplotObject <- ecodata::plot_finfish_traits(report = 'MidAtlantic', varName = 'age_maturity', n = 10) +
ggplot2::ggtitle('')
ggplotObject
try(downloadthis::download_this(
ggplotObject$data,
output_name = 'finfish_traitsMidAtlanticage_maturity',
output_extension = '.csv',
button_label = 'Download the data plotted above',
button_type = 'default',
has_icon = TRUE,
icon = 'fa fa-save',
class = 'hvr-sweep-to-left',
csv2 = F),
silent = TRUE)
```
<br>

**New England, age_maturity**
```{r plot_finfish_traitsNewEnglandage_maturity}
# Plot indicator
ggplotObject <- ecodata::plot_finfish_traits(report = 'NewEngland', varName = 'age_maturity', n = 10) +
ggplot2::ggtitle('')
ggplotObject
try(downloadthis::download_this(
ggplotObject$data,
output_name = 'finfish_traitsNewEnglandage_maturity',
output_extension = '.csv',
button_label = 'Download the data plotted above',
button_type = 'default',
has_icon = TRUE,
icon = 'fa fa-save',
class = 'hvr-sweep-to-left',
csv2 = F),
silent = TRUE)
```
<br>

**Mid-Atlantic, length_maturity**
```{r plot_finfish_traitsMidAtlanticlength_maturity}
# Plot indicator
ggplotObject <- ecodata::plot_finfish_traits(report = 'MidAtlantic', varName = 'length_maturity', n = 10) +
ggplot2::ggtitle('')
ggplotObject
try(downloadthis::download_this(
ggplotObject$data,
output_name = 'finfish_traitsMidAtlanticlength_maturity',
output_extension = '.csv',
button_label = 'Download the data plotted above',
button_type = 'default',
has_icon = TRUE,
icon = 'fa fa-save',
class = 'hvr-sweep-to-left',
csv2 = F),
silent = TRUE)
```
<br>

**New England, length_maturity**
```{r plot_finfish_traitsNewEnglandlength_maturity}
# Plot indicator
ggplotObject <- ecodata::plot_finfish_traits(report = 'NewEngland', varName = 'length_maturity', n = 10) +
ggplot2::ggtitle('')
ggplotObject
try(downloadthis::download_this(
ggplotObject$data,
output_name = 'finfish_traitsNewEnglandlength_maturity',
output_extension = '.csv',
button_label = 'Download the data plotted above',
button_type = 'default',
has_icon = TRUE,
icon = 'fa fa-save',
class = 'hvr-sweep-to-left',
csv2 = F),
silent = TRUE)
```
<br>

**Mid-Atlantic, fecundity**
```{r plot_finfish_traitsMidAtlanticfecundity}
# Plot indicator
ggplotObject <- ecodata::plot_finfish_traits(report = 'MidAtlantic', varName = 'fecundity', n = 10) +
ggplot2::ggtitle('')
ggplotObject
try(downloadthis::download_this(
ggplotObject$data,
output_name = 'finfish_traitsMidAtlanticfecundity',
output_extension = '.csv',
button_label = 'Download the data plotted above',
button_type = 'default',
has_icon = TRUE,
icon = 'fa fa-save',
class = 'hvr-sweep-to-left',
csv2 = F),
silent = TRUE)
```
<br>

**New England, fecundity**
```{r plot_finfish_traitsNewEnglandfecundity}
# Plot indicator
ggplotObject <- ecodata::plot_finfish_traits(report = 'NewEngland', varName = 'fecundity', n = 10) +
ggplot2::ggtitle('')
ggplotObject
try(downloadthis::download_this(
ggplotObject$data,
output_name = 'finfish_traitsNewEnglandfecundity',
output_extension = '.csv',
button_label = 'Download the data plotted above',
button_type = 'default',
has_icon = TRUE,
icon = 'fa fa-save',
class = 'hvr-sweep-to-left',
csv2 = F),
silent = TRUE)
```
<br>

**Mid-Atlantic, l_inf**
```{r plot_finfish_traitsMidAtlanticl_inf}
# Plot indicator
ggplotObject <- ecodata::plot_finfish_traits(report = 'MidAtlantic', varName = 'l_inf', n = 10) +
ggplot2::ggtitle('')
ggplotObject
try(downloadthis::download_this(
ggplotObject$data,
output_name = 'finfish_traitsMidAtlanticl_inf',
output_extension = '.csv',
button_label = 'Download the data plotted above',
button_type = 'default',
has_icon = TRUE,
icon = 'fa fa-save',
class = 'hvr-sweep-to-left',
csv2 = F),
silent = TRUE)
```
<br>

**New England, l_inf**
```{r plot_finfish_traitsNewEnglandl_inf}
# Plot indicator
ggplotObject <- ecodata::plot_finfish_traits(report = 'NewEngland', varName = 'l_inf', n = 10) +
ggplot2::ggtitle('')
ggplotObject
try(downloadthis::download_this(
ggplotObject$data,
output_name = 'finfish_traitsNewEnglandl_inf',
output_extension = '.csv',
button_label = 'Download the data plotted above',
button_type = 'default',
has_icon = TRUE,
icon = 'fa fa-save',
class = 'hvr-sweep-to-left',
csv2 = F),
silent = TRUE)
```
<br>

**Mid-Atlantic, k**
```{r plot_finfish_traitsMidAtlantick}
# Plot indicator
ggplotObject <- ecodata::plot_finfish_traits(report = 'MidAtlantic', varName = 'k', n = 10) +
ggplot2::ggtitle('')
ggplotObject
try(downloadthis::download_this(
ggplotObject$data,
output_name = 'finfish_traitsMidAtlantick',
output_extension = '.csv',
button_label = 'Download the data plotted above',
button_type = 'default',
has_icon = TRUE,
icon = 'fa fa-save',
class = 'hvr-sweep-to-left',
csv2 = F),
silent = TRUE)
```
<br>

**New England, k**
```{r plot_finfish_traitsNewEnglandk}
# Plot indicator
ggplotObject <- ecodata::plot_finfish_traits(report = 'NewEngland', varName = 'k', n = 10) +
ggplot2::ggtitle('')
ggplotObject
try(downloadthis::download_this(
ggplotObject$data,
output_name = 'finfish_traitsNewEnglandk',
output_extension = '.csv',
button_label = 'Download the data plotted above',
button_type = 'default',
has_icon = TRUE,
icon = 'fa fa-save',
class = 'hvr-sweep-to-left',
csv2 = F),
silent = TRUE)
```
<br>

**Mid-Atlantic, max_obs_length**
```{r plot_finfish_traitsMidAtlanticmax_obs_length}
# Plot indicator
ggplotObject <- ecodata::plot_finfish_traits(report = 'MidAtlantic', varName = 'max_obs_length', n = 10) +
ggplot2::ggtitle('')
ggplotObject
try(downloadthis::download_this(
ggplotObject$data,
output_name = 'finfish_traitsMidAtlanticmax_obs_length',
output_extension = '.csv',
button_label = 'Download the data plotted above',
button_type = 'default',
has_icon = TRUE,
icon = 'fa fa-save',
class = 'hvr-sweep-to-left',
csv2 = F),
silent = TRUE)
```
<br>

**New England, max_obs_length**
```{r plot_finfish_traitsNewEnglandmax_obs_length}
# Plot indicator
ggplotObject <- ecodata::plot_finfish_traits(report = 'NewEngland', varName = 'max_obs_length', n = 10) +
ggplot2::ggtitle('')
ggplotObject
try(downloadthis::download_this(
ggplotObject$data,
output_name = 'finfish_traitsNewEnglandmax_obs_length',
output_extension = '.csv',
button_label = 'Download the data plotted above',
button_type = 'default',
has_icon = TRUE,
icon = 'fa fa-save',
class = 'hvr-sweep-to-left',
csv2 = F),
silent = TRUE)
```
<br>

**Mid-Atlantic, PC1**
```{r plot_finfish_traitsMidAtlanticPC1}
# Plot indicator
ggplotObject <- ecodata::plot_finfish_traits(report = 'MidAtlantic', varName = 'PC1', n = 10) +
ggplot2::ggtitle('')
ggplotObject
try(downloadthis::download_this(
ggplotObject$data,
output_name = 'finfish_traitsMidAtlanticPC1',
output_extension = '.csv',
button_label = 'Download the data plotted above',
button_type = 'default',
has_icon = TRUE,
icon = 'fa fa-save',
class = 'hvr-sweep-to-left',
csv2 = F),
silent = TRUE)
```
<br>

**New England, PC1**
```{r plot_finfish_traitsNewEnglandPC1}
# Plot indicator
ggplotObject <- ecodata::plot_finfish_traits(report = 'NewEngland', varName = 'PC1', n = 10) +
ggplot2::ggtitle('')
ggplotObject
try(downloadthis::download_this(
ggplotObject$data,
output_name = 'finfish_traitsNewEnglandPC1',
output_extension = '.csv',
button_label = 'Download the data plotted above',
button_type = 'default',
has_icon = TRUE,
icon = 'fa fa-save',
class = 'hvr-sweep-to-left',
csv2 = F),
silent = TRUE)
```
<br>

**Mid-Atlantic, PC2**
```{r plot_finfish_traitsMidAtlanticPC2}
# Plot indicator
ggplotObject <- ecodata::plot_finfish_traits(report = 'MidAtlantic', varName = 'PC2', n = 10) +
ggplot2::ggtitle('')
ggplotObject
try(downloadthis::download_this(
ggplotObject$data,
output_name = 'finfish_traitsMidAtlanticPC2',
output_extension = '.csv',
button_label = 'Download the data plotted above',
button_type = 'default',
has_icon = TRUE,
icon = 'fa fa-save',
class = 'hvr-sweep-to-left',
csv2 = F),
silent = TRUE)
```
<br>

**New England, PC2**
```{r plot_finfish_traitsNewEnglandPC2}
# Plot indicator
ggplotObject <- ecodata::plot_finfish_traits(report = 'NewEngland', varName = 'PC2', n = 10) +
ggplot2::ggtitle('')
ggplotObject
try(downloadthis::download_this(
ggplotObject$data,
output_name = 'finfish_traitsNewEnglandPC2',
output_extension = '.csv',
button_label = 'Download the data plotted above',
button_type = 'default',
has_icon = TRUE,
icon = 'fa fa-save',
class = 'hvr-sweep-to-left',
csv2 = F),
silent = TRUE)
```
<br>

**Mid-Atlantic, PC3**
```{r plot_finfish_traitsMidAtlanticPC3}
# Plot indicator
ggplotObject <- ecodata::plot_finfish_traits(report = 'MidAtlantic', varName = 'PC3', n = 10) +
ggplot2::ggtitle('')
ggplotObject
try(downloadthis::download_this(
ggplotObject$data,
output_name = 'finfish_traitsMidAtlanticPC3',
output_extension = '.csv',
button_label = 'Download the data plotted above',
button_type = 'default',
has_icon = TRUE,
icon = 'fa fa-save',
class = 'hvr-sweep-to-left',
csv2 = F),
silent = TRUE)
```
<br>

**New England, PC3**
```{r plot_finfish_traitsNewEnglandPC3}
# Plot indicator
ggplotObject <- ecodata::plot_finfish_traits(report = 'NewEngland', varName = 'PC3', n = 10) +
ggplot2::ggtitle('')
ggplotObject
try(downloadthis::download_this(
ggplotObject$data,
output_name = 'finfish_traitsNewEnglandPC3',
output_extension = '.csv',
button_label = 'Download the data plotted above',
button_type = 'default',
has_icon = TRUE,
icon = 'fa fa-save',
class = 'hvr-sweep-to-left',
csv2 = F),
silent = TRUE)
```
<br>

## Implications
In general, there appears to be equivocal evidence for functional redundancy in traits at both the scale of the NE Shelf and within particular EPU’s. For instance, for trait ordination axes related to the pace of life and fecundity, there was stability in trait distributions for species captured during the fall survey. However, in the spring survey there was evidence for long-term shifts towards slower life history strategies (larger body size, lower fecundity, fewer offspring). 

At the EPU scale, there was little evidence for long-term linear trends in trait distributions in the Gulf of Maine and Scotian Shelf finfish communities, suggesting a high level of functional redundancy and stability in the face of dramatic shifts in oceanographic conditions (e.g. increasing temperatures) and other anthropogenic impacts. Conversely, there was evidence for long term change in trait distributions in the mid-Atlantic Bight and Georges Bank. For instance, the fall finfish community on George’s Bank and the spring finfish community in the mid-Atlantic Bight are both showing long-term shifts towards slower life history strategies with lower fecundity. These changes suggest that more research across the finfish community in these regions is needed to understand the ecological and economic ramifications of such change.

## Indicator statistics 
Spatial scale: By EPU and full shelf

Temporal scale: 1963-2023, Spring (January-June) & Fall (July-December)

**Variable definitions**

```{r vars_finfish_traits}
# Pull all var names
vars <- ecodata::finfish_traits |>
   dplyr::select(Var, Units) |>
   dplyr::distinct()

DT::datatable(vars)
```
## Get the data

**Point of contact**: Bart DiFiore (bartholomew.p.difiore@mass.gov)

**ecodata dataset**: `ecodata::finfish_traits`

**Tech Doc link**: <https://noaa-edab.github.io/tech-doc/finfish_traits.html>

### Public Availability

Source data are publicly available.

