# Harmful Algal Blooms

**Description**: These data represent annual estimated abundance of Alexandrium catanella cysts in the Gulf of Maine and the presence of PSP toxins in blue mussels at coastal sites in the Gulf of Maine (MA, NH, ME), and shellfishery closures (MA).

**Indicator family**: 

- [X] Lower trophic levels


**Contributor(s)**: Yizhen Li, NOAA/NOS NCCOS Stressor Detection and Impacts Division, HAB Forecasting Branch, Silver Spring MD

**Affiliations**: ?

```{r echo=FALSE}
knitr::opts_chunk$set(echo = F)
library(ecodata)
```
## Introduction to Indicator
Alexandrium cysts in sediments of the Gulf of Maine have been monitored through a cooperative effort of NOAA, WHOI, and other partners for over twenty years. Sampling methods are described in Anderson et al. 2005. In the annual survey cruises, samples are obtained with a Craib corer, and Alexandrium cysts are counted from the top 1- cm of sediment layer. Results are extrapolated to estimate overall cyst abundance in the eastern, western, and entire Gulf of Maine.

These data represent the presence of PSP toxins in blue mussels (Mytilis edulis) sampled at coastal sites in Massachusetts (1972-2022), New Hampshire (2000-2022), and Maine (2005 to 2019). Results are summarized on an annual basis for each state. Variables include total number of samples, number of samples above and below a designated threshold (44ug STX equivalent / 100g tissue), and percentage of samples above the threshold. These data were provided by Yizhen Li, NOAA/NOS NCCOS Stressor Detection and Impacts Division, HAB Forecasting Branch, Silver Spring, MD. Original data for Maine were collected by Maine Department of Marine Resources, which tests coastal shellfish areas for biotoxins weekly, annually beginning in March and going through October or later when necessary. Original data for New Hampshire were collected by NH Department of Environmental Services, and original data for Massachusetts were collected by Massachusetts Division of Marine Fisheries.

## Key Results and Visualizations
Visualizations of the data include an annual time series plot of estimated cyst abundance, and a time series of maps depicting the spatial and temporal variability in the Gulf of Maine.  Results are plotted as estimated total numbers of cells (10 to the 16th power) in Eastern Gulf of Maine (east of Penobscot Bay), Western Gulf of Maine (west of Penobscot Bay), Bay of Fundy (2003-2013 only), and entire Gulf of Maine.

A line graph depicts the percentage of samples that exceed the designated threshold (44ug STX equivalent / 100g tissue), in the samples that had some level of STX detected. The percentages may be affected by protocols on where and when to sample – e.g. sampling may be more intense during bloom events, and protocols may differ between the three states (MA, NH, ME).
A three-tier column graph illustrates three different metrics of shellfish bed closures in Massachusetts waters due to PSP, 1990-2022. Annual metrics include number of sites closed, mean number of days closed per site, and total closed area (acres). Bloom events in the early 1990s, 2005, and 2009 are evident in the closure metrics.

### MidAtlantic

```{r plot_habsMidAtlanticAlexandrium}
# Plot indicator
ggplotObject <- ecodata::plot_habs(report= 'MidAtlantic', varName= 'Alexandrium')
print(ggplotObject)
```

```{r plot_habsMidAtlanticPSP}
# Plot indicator
ggplotObject <- ecodata::plot_habs(report= 'MidAtlantic', varName= 'PSP')
print(ggplotObject)
```

### NewEngland

```{r plot_habsNewEnglandAlexandrium}
# Plot indicator
ggplotObject <- ecodata::plot_habs(report= 'NewEngland', varName= 'Alexandrium')
print(ggplotObject)
```

```{r plot_habsNewEnglandPSP}
# Plot indicator
ggplotObject <- ecodata::plot_habs(report= 'NewEngland', varName= 'PSP')
print(ggplotObject)
```


## Indicator statistics 
Spatial scale: Alexandrium results are reported for the entire Gulf of Maine, and also by sub-regions: Eastern Gulf of Maine (east of Penobscot Bay), Western Gulf of Maine (west of Penobscot Bay), and Bay of Fundy (2003-2013 only). PSP results are based on data from individual sites in the Gulf of Maine, aggregated by year and state (MA, NH, ME).

Temporal scale: Alexandrium results are aggregated annually, and reported for 2003 to 2021.  Cyst sampling is done in the late fall. PSP results are aggregated by year.

**Synthesis Theme**:

- [X] Regime Shifts


```{r autostats_habs}
# Either from Contributor or ecodata
```

## Implications
The annual fall surveys of Alexandrium cyst distribution and abundance in the Gulf of Maine provide data needed to predict bloom events in the following year. Other key variables in the predictive models include currents, nutrients, temperature, and salinity dynamics at multiple spatial and temporal scales (Anderson et al. 2005, Li et al. 2009 & 2020, McGillicuddy et al. 2011). After strong cyst deposition events in 2005 and 2009, the time series suggest lower overall cyst abundance through 2021. However, bloom events and shellfishery closures usually occur somewhere in any given year. Economic impacts to inshore shellfisheries (mussels, softshell clams, quahogs) can be substantial (Jin and Hoagland 2008). Impacts to offshore shellfisheries can occur as well, for example surf clam and ocean quahog on Georges Bank 1988-1990 (Anderson et al. 1997).

Alexandrium bloom events in the Gulf of Maine can result in Paralytic Shellfish Poisoning (PSP) toxins accumulating in shellfish and other species. Bloom events and shellfishery closures usually occur somewhere in any given year, but vary greatly between years and among local areas. The presence of PSP toxins in shellfish, and shellfishery closures have been used as metrics for assessing the severity of PSP outbreaks (Kleindinst et al. 2014). Economic impacts to inshore shellfisheries (mussels, softshell clams, quahogs) can be substantial (Jin and Hoagland 2008).

## Get the data

**Point of contact**: [Moe Nelson, david.moe.nelson@noaa.gov](mailto:Moe Nelson, david.moe.nelson@noaa.gov){.email}

**ecodata name**: `ecodata::habs`

**Variable definitions**

Alexandrium: 1) Year; Definition: calendar year; Units: yyyy. 
2) Name: Var; Definition: Gulf of Maine region (West, East, Bay of Fundy, All); Units: categories. 
3) Name: Value; Definition: Estimated cyst abundance; Units: numbers of cells * 10 to the 16th power)  PSP: 1) Year; Definition: calendar year; Units: yyyy. 
2) Name: State; Definition: MA, NH, or ME ; Units: categories. 3) Name: N_Rows; Definition: Number of sample events represented; Units: integer.  
4) Name: PSP_Exceed_Threshold_Pct; Definition: Percentage of samples exceeding PSP threshold; Units: decimal number, 0-100.

```{r vars_habs}
# Pull all var names
vars <- ecodata::habs |>
   dplyr::select(Var) |>
   dplyr::distinct()

DT::datatable(vars)
```
**Indicator Category**:

- [X] Published Methods
- [X] Database pull with analysis


## Public Availability

Source data are NOT publicly available.

## Accessibility and Contraints

Data were provided by consultation with Yizhen Li.  Data are also used in operational HAB forecast models, freely available to the public. Data as aggregated are not publicly available. Data can be acquired upon request.

**tech-doc link**
<https://noaa-edab.github.io/tech-doc/habs.html>

## References

