# Marine Heatwave

**Description**: Marine Heatwave

**Contributor(s)**: Vincent Saba
  
**Data steward**: Kimberly Bastille <kimberly.bastille@noaa.gov>
  
**Point of contact**: Vincent Saba <vincent.saba@noaa.gov>


## Impact Summary

Marine heatwaves measure not just temperature, but how long the ecosystem is subjected to the high temperature. They are driven by both atmospheric and oceanographic factors and can have dramatic impacts on marine ecosystems. Marine heatwaves are measured in terms of intensity (water temperature) and duration (the cumulative number of degree days) using satellite measurements of daily sea surface temperature. Plotted below are maximum intensity and cumulative intensity, which is intensity times duration. Here we define a marine heatwave as a warming event that lasts for five or more days with sea surface temperatures above the 90th percentile of the historical daily climatology (1982-2011). 

### Georges Bank

In 2020, Georges Bank experienced six distinct marine heatwaves in the spring, summer, and fall, which included the warmest event on record.  The strongest event began on July 8 and lasting 51 days. Relative to prior years, this marine heatwave ranked 1st on record in terms of maximum intensity (+4.33 °C above average on Aug 14) and 4th on record in terms of cumulative intensity (138 oC-days). Another relatively strong marine heatwave began on October 20 and also lasted 51 days with a peak warming on October 30 (+3.17 °C above average).   This event was the 8th strongest on record in terms of cumulative intensity (111 oC-days).  The strongest three marine heatwaves on record in Georges Bank occurred over the last five years (2015, 2016, 2020).
```{r , code = readLines("https://raw.githubusercontent.com/NOAA-EDAB/ecodata/master/chunk-scripts/LTL_MAB.Rmd-heatwave-gb.R"), eval = T, echo=F, fig.cap="Cumulative and maximum marine heatwave in Georges Bank", fig.align="center"}

```

```{r, fig.cap="Georges Bank Marine Heatwave events in 2020.",code = readLines("https://raw.githubusercontent.com/NOAA-EDAB/ecodata/master/chunk-scripts/LTL_MAB.Rmd-heatwave-year-gb.R"), eval = T, echo=F}

```
  
### Gulf of Maine

In 2020, the Gulf of Maine experienced five distinct marine heatwaves throughout all seasons with the strongest event beginning on June 17 and lasting 71 days. Relative to prior years, this marine heatwave ranked 7th on record in terms of maximum intensity (+4.02 °C above average on August 2) and 5th on record in terms of cumulative intensity (206 °C-days ) . 
```{r , code = readLines("https://raw.githubusercontent.com/NOAA-EDAB/ecodata/master/chunk-scripts/LTL_NE.Rmd-heatwave-gom.R"), eval = T, echo=F, fig.cap="Cumulative and maximum marine heatwave in the Gulf of Maine", fig.align="center"}

```

```{r, fig.cap="Gulf of Maine Marine Heatwave events in 2020.",code = readLines("https://raw.githubusercontent.com/NOAA-EDAB/ecodata/master/chunk-scripts/LTL_NE.Rmd-heatwave-year-gom.R"), eval = T, echo=F}

```

### Mid-Atlantic Bight

In 2020, the Middle Atlantic Bight experienced eight distinct marine heatwaves with the strongest event beginning on July 18 and lasting 18 days. Relative to prior years, this marine heatwave ranked 30th on record in terms of maximum intensity (+2.61 °C above average on July 28) and 26th on record in terms of cumulative intensity (40 °C-days ).  The region was not as warm as in recent years but is still experiencing more frequent marine heatwaves over the last decade, including this year, compared to the historical period from 1980’s to early 2000’s. 

```{r , code = readLines("https://raw.githubusercontent.com/NOAA-EDAB/ecodata/master/chunk-scripts/LTL_MAB.Rmd-heatwave.R"), eval = T, echo=F, fig.cap="Cumulative and maximum marine heatwave in the Mid-Atlantic", fig.align="center"}

```

```{r, fig.cap="Mid-Atlantic Marine Heatwave events in 2020.",code = readLines("https://raw.githubusercontent.com/NOAA-EDAB/ecodata/master/chunk-scripts/LTL_MAB.Rmd-heatwave-year.R"), eval = T, echo=F}

```

### Raw Data 
The raw data is available in the [`ecodata` package](https://noaa-edab.github.io/ecodata/landing_page). 

```{r}
library(DT)
datatable(ecodata::heatwave, options = list(pageLength = 10))
```


## Methods
Marine heatwave analysis for Georges Bank, Gulf of Maine, and the Middle Atlantic Bight according to the definition in @hobday2016.  

### Data sources
NOAA high-res OISST (daily, 25-km, 1982-2019)   https://www.esrl.noaa.gov/psd/cgi-bin/db_search/DBListFiles.pl?did=132&tid=79458&vid=2423

### Data extraction 
Each yearly file (global) was downloaded, concatenated into a single netcdf file using nco (Unix), and then cropped to the USNES region using Ferret.  Each EPU's time-series of SST was averaged using .shp file boundaries for the MAB, GB, and GOM (also done in Ferret) and saved to the three .csv files.

### Data analysis
The marine heatwave metrics Maximum Intensity [deg. C] and Cumulative Intensity [deg. C x days] are calculated using NOAA OISST daily sea surface temperature data (25-km resolution) from January 1982 to December 2019.  The heatwaves are calculated based on the algorithms in Hobday et al. 2016 and by using a climatology of 1982-2011.  These metrics were run R using https://robwschlegel.github.io/heatwaveR/

### Data processing 
Marine Heatwave data were formatted for inclusion in the `ecodata` R package using this [R code](https://github.com/NOAA-EDAB/ecodata/blob/master/data-raw/get_heatwave.R).


