# Atlantic Highly Migratory Species Stock Status {#hms_stock_status}

```{r echo=FALSE}
knitr::opts_chunk$set(echo = F)
library(ecodata)
library(downloadthis)
```

*Last updated July 23, 2026*

**Description**: Summary of stock assessment results for each assessed HMS species as of December 2023.

**Contributor(s)**: Jennifer Cudney, Heather Baertlein, Jackie Wilson, George Silva

**Affiliations**: NEFSC

**Indicator Family**: 

- [X] Megafauna


**Indicator Category**:

- [X] Syntheses of published information


**Synthesis Theme**:

- [X] Regime Shifts


```{r download_hms_stock_status}
# Download dataset from ecodata
try(downloadthis::download_this(
ecodata::hms_stock_status,
output_name = 'hms_stock_status',
output_extension = '.csv',
button_label = 'Download dataset from `ecodata`',
button_type = 'default',
has_icon = TRUE,
icon = 'fa fa-save',
class = 'hvr-sweep-to-left',
csv2 = F),
silent = TRUE)
```

## Introduction to Indicator
This section summarizes Highly Migratory Species Stock Status as it is presented in the most recent Atlantic Highly Migratory Species Stock Assessment and Fishery Evaluation (SAFE) report (2024-2025). Stock assessments for Atlantic HMS are completed through one of three processes. Most Atlantic HMS stocks that are managed internationally under the authority of the MSA and ATCA (tunas, billfish, swordfish, some pelagic sharks) are assessed by the International Commission for the Conservation of Atlantic Tunas’ Standing Committee on Research and Statistics (ICCAT SCRS). Sharks managed domestically under the MSA are assessed through the SouthEast Data Assessment and Review (SEDAR) process. Sharks may also be assessed through an independent stock assessment is completed by a third party and approved for use in management by NOAA Fisheries (“external assessments”). Most HMS assessments are completed through either the ICCAT or SEDAR process.  Separate tables summarize thresholds and stock status for all species or stocks assessed through these pathways (and for which the results have been accepted for use in management), and provide information on the timing of upcoming assessments.

## Key Results and Visualizations
The domestic and international status of 28 species or stocks are described in the most recent HMS SAFE report. 

Based on the thresholds identifying whether a stock is considered “overfished” in the HMS FMP, 13 species or stocks are not considered overfished, 10 species or stocks are considered overfished, and 5 species or stocks have unknown status. 

Based on the thresholds identifying whether a stock is experiencing “overfishing” in the HMS FMP, overfishing is not occurring (or not likely occurring) for 19 species or stocks, overfishing is occurring for 5 species or stocks, and the status of 4 species or stocks is unknown.

**Mid-Atlantic**
```{r plot_hms_stock_statusMidAtlantic}
# Plot indicator
ggplotObject <- ecodata::plot_hms_stock_status(report = 'MidAtlantic') +
ggplot2::ggtitle('')
ggplotObject
try(downloadthis::download_this(
ggplotObject$data,
output_name = 'hms_stock_statusMidAtlantic',
output_extension = '.csv',
button_label = 'Download the data plotted above',
button_type = 'default',
has_icon = TRUE,
icon = 'fa fa-save',
class = 'hvr-sweep-to-left',
csv2 = F),
silent = TRUE)
```
<br>

**New England**
```{r plot_hms_stock_statusNewEngland}
# Plot indicator
ggplotObject <- ecodata::plot_hms_stock_status(report = 'NewEngland') +
ggplot2::ggtitle('')
ggplotObject
try(downloadthis::download_this(
ggplotObject$data,
output_name = 'hms_stock_statusNewEngland',
output_extension = '.csv',
button_label = 'Download the data plotted above',
button_type = 'default',
has_icon = TRUE,
icon = 'fa fa-save',
class = 'hvr-sweep-to-left',
csv2 = F),
silent = TRUE)
```
<br>

## Implications
This summarizes information in the HMS SAFE Report, which itself is intended to provide a high-level summary of stock status information. Specific implications of stock status trends through time are available in the latest assessments.

## Indicator statistics 
Spatial scale: Species-specific; depends on the extent of the managed stock. Can be regional (e.g., Gulf of America or Atlantic blacktip shark) or could be basin-wide (e.g., North Atlantic swordfish, Western Atlantic bluefin)

Temporal scale: Stock status as of December 2025. Status is stock-specific with regards to temporal nature of thresholds, estimated rebuilding times, and data analyzed for stock assessments.

**Variable definitions**

```{r vars_hms_stock_status}
# Pull all var names
vars <- ecodata::hms_stock_status |>
   dplyr::select(Var) |>
   dplyr::distinct()

DT::datatable(vars)
```
## Get the data

**Point of contact**: Jennifer Cudney (jennifer.cudney@noaa.gov)

**ecodata dataset**: `ecodata::hms_stock_status`

**Tech Doc link**: <https://noaa-edab.github.io/tech-doc/hms_stock_status.html>

### Public Availability

Source data are publicly available.

