# NE Shelf Annual Sea Surface Temperature (SST) {#long_term_sst}

```{r echo=FALSE}
knitr::opts_chunk$set(echo = F)
library(ecodata)
library(downloadthis)
```

*Last updated July 23, 2026*

**Description**: Average annual sea-surface temperatures from the NOAA extended reconstructed sea surface temperature data set (ERSST V5) on the Northeast Continental Shelf.

**Contributor(s)**: Kevin Friedland, Brandon Beltz

**Affiliations**: NEFSC

**Indicator Family**: 

- [X] Oceanographic


**Indicator Category**:

- [X] Database pull


**Synthesis Theme**:

- [X] Multiple System Drivers


```{r download_long_term_sst}
# Download dataset from ecodata
try(downloadthis::download_this(
ecodata::long_term_sst,
output_name = 'long_term_sst',
output_extension = '.csv',
button_label = 'Download dataset from `ecodata`',
button_type = 'default',
has_icon = TRUE,
icon = 'fa fa-save',
class = 'hvr-sweep-to-left',
csv2 = F),
silent = TRUE)
```

## Introduction to Indicator
The data presented here are average annual sea-surface temperatures from the NOAA extended reconstructed sea surface temperature data set (ERSST V5) on the Northeast Continental Shelf.

## Key Results and Visualizations
Since the 1860’s, the Northeast US shelf sea surface temperature (SST) has exhibited an overall warming trend, with the past decade measuring well above the long term average.

**Mid-Atlantic**
```{r plot_long_term_sstMidAtlantic}
# Plot indicator
ggplotObject <- ecodata::plot_long_term_sst(report = 'MidAtlantic', n = 10) +
ggplot2::ggtitle('')
ggplotObject
try(downloadthis::download_this(
ggplotObject$data,
output_name = 'long_term_sstMidAtlantic',
output_extension = '.csv',
button_label = 'Download the data plotted above',
button_type = 'default',
has_icon = TRUE,
icon = 'fa fa-save',
class = 'hvr-sweep-to-left',
csv2 = F),
silent = TRUE)
```
<br>

## Implications
Long term SST demonstrates that the annual average sea surface temperatures observed during the most recent ten years are well above the historical average dating back to the 1860s.

## Indicator statistics 
Spatial scale: Full shelf

Temporal scale: Annual

**Variable definitions**

```{r vars_long_term_sst}
# Pull all var names
vars <- ecodata::long_term_sst |>
   dplyr::select(Var, Units) |>
   dplyr::distinct()

DT::datatable(vars)
```
## Get the data

**Point of contact**: Brandon.Beltz@noaa.gov

**ecodata dataset**: `ecodata::long_term_sst`

**Tech Doc link**: <https://noaa-edab.github.io/tech-doc/long_term_sst.html>

### Public Availability

Source data are publicly available.

