# NE Shelf Annual Sea Surface Temperature (SST) {#long_term_sst}

**Description**: Average annual sea-surface temperatures from the NOAA extended reconstructed sea surface temperature data set (ERSST V5) on the Northeast Continental Shelf.

**Indicator family**: 

- [X] Oceanographic


**Contributor(s)**: Kevin Friedland, Brandon Beltz

**Affiliations**: NEFSC

```{r echo=FALSE}
knitr::opts_chunk$set(echo = F)
library(ecodata)
```
## Introduction to Indicator
The data presented here are average annual sea-surface temperatures from the NOAA extended reconstructed sea surface temperature data set (ERSST V5) on the Northeast Continental Shelf.

## Key Results and Visualizations
Since the 1860’s, the Northeast US shelf sea surface temperature (SST) has exhibited an overall warming trend, with the past decade measuring well above the long term average.

```{r plot_long_term_sstMAB}
# Plot indicator
ggplotObject <- ecodata::plot_long_term_sst(report='MidAtlantic')
ggplotObject
```


## Indicator statistics 
Spatial scale: Full shelf

Temporal scale: Annual

**Synthesis Theme**:

- [X] Multiple System Drivers


```{r autostats_long_term_sst}
# Either from Contributor or ecodata
```

## Implications
Long term SST demonstrates that the annual average sea surface temperatures observed during the most recent ten years are well above the historical average dating back to the 1860s.

## Get the data

**Point of contact**: [Brandon.Beltz@noaa.gov](mailto:Brandon.Beltz@noaa.gov){.email}

**ecodata name**: `ecodata::long_term_sst`

**Variable definitions**

long-term sst  in  units of degreesC

```{r vars_long_term_sst}
# Pull all var names
vars <- ecodata::long_term_sst |>
   dplyr::select(Var, Units) |>
   dplyr::distinct()

DT::datatable(vars)
```
**Indicator Category**:

- [X] Database pull


## Public Availability

Source data are publicly available.

## Accessibility and Constraints

_No response_

**tech-doc link**
<https://noaa-edab.github.io/tech-doc/long_term_sst.html>

