# Inshore Survey (Mid Atlantic) {#mab_inshore_survey}

```{r echo=FALSE}
knitr::opts_chunk$set(echo = F)
library(ecodata)
library(downloadthis)
```

*Last updated July 23, 2026*

**Description**: Biomass time series for aggregate species groups from the NEAMAP inshore bottom trawl surveys conducted from Cape Hatteras to Rhode Island.

**Contributor(s)**: James Gartland, Matt Camisa, Rebecca Peters, Sean Lucey

**Affiliations**: VIMS, Maine

**Indicator Family**: 

- [X] Megafauna


**Indicator Category**:

- [X] Syntheses of published information
- [X] Database pull


**Synthesis Theme**:

- [X] Multiple System Drivers


```{r download_mab_inshore_survey}
# Download dataset from ecodata
try(downloadthis::download_this(
ecodata::mab_inshore_survey,
output_name = 'mab_inshore_survey',
output_extension = '.csv',
button_label = 'Download dataset from `ecodata`',
button_type = 'default',
has_icon = TRUE,
icon = 'fa fa-save',
class = 'hvr-sweep-to-left',
csv2 = F),
silent = TRUE)
```

## Introduction to Indicator
Indicators from these inshore surveys are analogous to those produced by the NEFSC trawl survey in the [aggregate survey biomass](https://noaa-edab.github.io/catalog/aggregate_biomass.html) indicator dataset.

## Key Results and Visualizations
Each survey shows trends by aggregate group.

**Mid-Atlantic**
```{r plot_mab_inshore_surveyMidAtlantic}
# Plot indicator
ggplotObject <- ecodata::plot_mab_inshore_survey(report = 'MidAtlantic', n = 10) +
ggplot2::ggtitle('')
ggplotObject
try(downloadthis::download_this(
ggplotObject$data,
output_name = 'mab_inshore_surveyMidAtlantic',
output_extension = '.csv',
button_label = 'Download the data plotted above',
button_type = 'default',
has_icon = TRUE,
icon = 'fa fa-save',
class = 'hvr-sweep-to-left',
csv2 = F),
silent = TRUE)
```
<br>

**New England**
```{r plot_mab_inshore_surveyNewEngland}
# Plot indicator
ggplotObject <- ecodata::plot_mab_inshore_survey(report = 'NewEngland', n = 10) +
ggplot2::ggtitle('')
ggplotObject
try(downloadthis::download_this(
ggplotObject$data,
output_name = 'mab_inshore_surveyNewEngland',
output_extension = '.csv',
button_label = 'Download the data plotted above',
button_type = 'default',
has_icon = TRUE,
icon = 'fa fa-save',
class = 'hvr-sweep-to-left',
csv2 = F),
silent = TRUE)
```
<br>

## Implications
The trends in the NEAMAP inshore survey vary compared to the NEFSC survey. This is driven by different species availability to surveys in time and space as well as the surveys sampling different habitats. Nearshore habitats are important to fish and fisheries so we monitor them as as well as trends from the larger EPUs.

## Indicator statistics 
Spatial scale: Nearshore regions of the MAB

Temporal scale: Spring and Fall

**Variable definitions**

```{r vars_mab_inshore_survey}
# Pull all var names
vars <- ecodata::mab_inshore_survey |>
   dplyr::select(Var) |>
   dplyr::distinct()

DT::datatable(vars)
```
## Get the data

**Point of contact**: James Gartland (NEAMAP), jgartlan@vims.edu; Rebecca Peters (ME/NH survey), rebecca.j.peters@maine.gov

**ecodata dataset**: `ecodata::mab_inshore_survey`

**Tech Doc link**: <https://noaa-edab.github.io/tech-doc/mab_inshore_survey.html>

### Public Availability

Source data are publicly available.

