# Inshore Survey (Massachusetts) {#mass_inshore_survey}

**Description**: Biomass time series for aggregate species groups from the Massachusetts inshore bottom trawl surveys conducted in Massachusetts state waters.

**Indicator family**: 

- [X] Megafauna


**Contributor(s)**: Sean Lucey

**Affiliations**: NEFSC

```{r echo=FALSE}
knitr::opts_chunk$set(echo = F)
library(ecodata)
```
## Introduction to Indicator
Indicators from this inshore survey are analogous to those produced by the NEFSC trawl survey in the [aggregate survey biomass](https://noaa-edab.github.io/catalog/aggregate_biomass.html) indicator dataset.

## Key Results and Visualizations
Each survey shows trends by aggregate group.

```{r plot_mass_inshore_surveyMidAtlantic}
# Plot indicator
ggplotObject <- ecodata::plot_mass_inshore_survey(report = 'MidAtlantic', n = 10)
ggplotObject
```

```{r plot_mass_inshore_surveyNewEngland}
# Plot indicator
ggplotObject <- ecodata::plot_mass_inshore_survey(report = 'NewEngland', n = 10)
ggplotObject
```


## Indicator statistics 
Spatial scale: Nearshore regions of Massachusetts

Temporal scale: Spring and Fall

**Synthesis Theme**:

- [X] Multiple System Drivers


```{r autostats_mass_inshore_survey}
# Either from Contributor or ecodata
```

## Implications
The trends in the Massachusetts inshore survey vary compared to the NEFSC survey. This is driven by different species availability to surveys in time and space as well as the surveys sampling different habitats. Nearshore habitats are important to fish and fisheries so we monitor them as as well as trends from the larger EPUs.

## Get the data

**Point of contact**: [EDAB, edab.data@noaa.gov](mailto:EDAB, edab.data@noaa.gov){.email}

**ecodata name**: `ecodata::mass_inshore_survey`

**Variable definitions**

See variable definitions for `aggregate_biomass`

```{r vars_mass_inshore_survey}
# Pull all var names
vars <- ecodata::mass_inshore_survey |>
   dplyr::select(Var, Units) |>
   dplyr::distinct()

DT::datatable(vars)
```
**Indicator Category**:

- [X] Syntheses of published information
- [X] Database pull


## Public Availability

Source data are publicly available.

## Accessibility and Constraints

_No response_

**tech-doc link**
<https://noaa-edab.github.io/tech-doc/mass_inshore_survey.html>

