# Survey Impacts from Offshore Wind Development {#osw_survey_impact}

**Description**: Offshore wind development is expected to have several impacts on federal and state surveys.

**Indicator family**: 

- [X] Social


**Contributor(s)**: Douglas Christel

**Affiliations**: GARFO

```{r echo=FALSE}
knitr::opts_chunk$set(echo = F)
library(ecodata)
```
## Introduction to Indicator
Proceed as though this were a short summary of a typical introduction section in a paper.

## Key Results and Visualizations
Proceed as though this were a short summary of a typical results section in a paper.


## Indicator statistics 
Spatial scale: full shelf

Temporal scale: Decadal

**Synthesis Theme**:

- [X] Multiple System Drivers


```{r autostats_osw_survey_impact}
# Either from Contributor or ecodata
```

## Implications
Proposed wind development areas interact with the region’s federal scientific surveys @friedland_spatial_2023. Scientific surveys are impacted by offshore wind in four ways:
 1)  Exclusion of NOAA Fisheries’ sampling platforms from the wind development area due to operational and safety limitations;
 2)  Impacts on the random-stratified statistical design that is the basis for scientific assessments, advice, and analyses; 
3) Alteration of benthic and pelagic habitats, and airspace in and around the wind energy development, requiring new designs and methods to sample new habitats; and, 
4) Reduced sampling productivity through navigation impacts of wind energy infrastructure on aerial and vessel survey operations. 

Increased vessel transit between stations may decrease data collections that are already limited by annual days-at-sea day allocations. The total survey area overlap ranges from 1-70% for all Greater Atlantic federal surveys. The Gulf of Maine Cooperative Research Bottom Longline Survey (41%) and the Shrimp Survey (70%) have the largest percent overlap with the draft Gulf of Maine Wind Energy Areas. The remaining surveys range from 1-16% overlap.  Individual survey strata have significant interaction with wind, including the sea scallop survey (up to 96% of individual strata) and the bottom trawl survey (BTS, up to 60% strata overlap). Additionally, up to 50% of the southern New England North Atlantic right whale survey’s area overlaps with proposed project areas and a region-wide survey mitigation program is underway @northeast_fisheries_science_center_us_fall_2022

## Get the data

**Point of contact**: [douglas.christel@noaa.gov](mailto:douglas.christel@noaa.gov){.email}

**ecodata name**: No dataset

**Variable definitions**

No Data


No Data

**Indicator Category**:

- [X] Syntheses of published information


## Public Availability

Source data are publicly available.

## Accessibility and Constraints

_No response_

