# Copepod Preyfield Energy Density {#preyfield_energy}

```{r echo=FALSE}
knitr::opts_chunk$set(echo = F)
library(ecodata)
library(downloadthis)
```

*Last updated July 23, 2026*

**Description**: These data show the estimated average preyfield energy available to foraging North Atlantic Right Whales on the shelf waters of the US northeast coast.

**Contributor(s)**: Gregory Ellis, Chris Orphanides, Jordan Katz, Harvey Walsh

**Affiliations**: NEFSC

**Indicator Family**: 

- [x] Lower trophic levels


**Indicator Category**:

- [x] Database pull with analysis


**Synthesis Theme**:

- [x] Multiple System Drivers


```{r download_preyfield_energy}
# Download dataset from ecodata
try(downloadthis::download_this(
ecodata::preyfield_energy,
output_name = 'preyfield_energy',
output_extension = '.csv',
button_label = 'Download dataset from `ecodata`',
button_type = 'default',
has_icon = TRUE,
icon = 'fa fa-save',
class = 'hvr-sweep-to-left',
csv2 = F),
silent = TRUE)
```

## Introduction to Indicator
The shelf waters of the northwest Atlantic are a principal feeding area of North Atlantic Right Whales. This visualization product uses interpolated abundance data from the forty year Ecomon zooplankton sampling database along with species specific per-animal energy estimates for three main NARW dietary components (Calanus finmarchicus, Centropages typicus, and Pseudocalanus spp) to show average monthly prey field energy levels available to these animals off the northeast coast.

## Key Results and Visualizations
The data show that available energy varies strongly between seasons and that maximum levels broadly coincide with periods of maximum habitat utilization by Right Whales.

## Implications
The product presented here provides a background for examining both long-term and interannual changes to prey availability for North Atlantic Right Whales as well as assisting in targeting limited sampling resources for future assessments.

## Indicator statistics 
Spatial scale: Shelf waters from Cape Hatteras to the Canadian Maritimes

Temporal scale: Maps are resolved on a monthly scale based on the average abundances observed across the entire Ecomon sampling period (1980-present). They are neither representative nor predictive of any specific year and should not be treated as such.

**Variable definitions**

```{r vars_preyfield_energy}
# Pull all var names
vars <- ecodata::preyfield_energy |>
   dplyr::select(Var) |>
   dplyr::distinct()

DT::datatable(vars)
```
## Get the data

**Point of contact**: Gregory Ellis (gregory.ellis@noaa.gov)

**ecodata dataset**: `ecodata::preyfield_energy`

**Tech Doc link**: <https://noaa-edab.github.io/tech-doc/preyfield_energy.html>

### Public Availability

Source data are publicly available.

