# Fish Productivity Indicators {#productivity_anomaly}

**Description**: Amount of small fish produced per large fish biomass over time

**Indicator family**: 

- [X] Megafauna


**Contributor(s)**: Sarah Gaichas, Andy Beet, Kimberly Bastille, Sean Lucey, and Charles Perretti.

**Affiliations**: NEFSC

```{r echo=FALSE}
knitr::opts_chunk$set(echo = F)
library(ecodata)
```
## Introduction to Indicator
The amount of young fish produced by a population tells us both about the health of individual populations and about the productivity throughout the ecosystem when we look across multiple stocks. This has implications for both managing fisheries (lower productivity often leads to lower harvest) and for other components of the ecosystem. 

These indicators are based on the work of Perretti [@perretti_regime_2017]. We updated the data used in that analysis to describe patterns of aggregate fish productivity in the Mid-Atlantic, Georges Bank, and Gulf of Maine by evaluating changes in reproductive output relative to adult population size across multiple stocks. Both survey information and stock assessment information is used in separate indicators.

## Key Results and Visualizations
The small fish per large fish anomaly indicator, derived from NEFSC bottom trawl survey data, shows that productivity has been declining in the Mid-Atlantic region since 2010.  A similar analysis based on stock assessment model outputs (recruitment per spawning stock biomass anomaly) for stocks primarily inhabiting the Mid-Atlantic region also shows a decline in productivity. The indicators show great variability in the Gulf of Maine and Georges Bank, with both regions being below average for much of the past decade.

### MidAtlantic

```{r plot_productivity_anomalyMidAtlanticanomalyMAB}
# Plot indicator
ggplotObject <- ecodata::plot_productivity_anomaly(report= 'MidAtlantic', varName= 'anomaly' ,EPU= 'MAB')
ggplotObject
```

```{r plot_productivity_anomalyMidAtlanticassessmentMAB}
# Plot indicator
ggplotObject <- ecodata::plot_productivity_anomaly(report= 'MidAtlantic', varName= 'assessment' ,EPU= 'MAB')
ggplotObject
```

### NewEngland

```{r plot_productivity_anomalyNewEnglandanomalyGB}
# Plot indicator
ggplotObject <- ecodata::plot_productivity_anomaly(report= 'NewEngland', varName= 'anomaly' ,EPU= 'GB')
ggplotObject
```

```{r plot_productivity_anomalyNewEnglandanomalyGOM}
# Plot indicator
ggplotObject <- ecodata::plot_productivity_anomaly(report= 'NewEngland', varName= 'anomaly' ,EPU= 'GOM')
ggplotObject
```

```{r plot_productivity_anomalyNewEnglandassessmentGB}
# Plot indicator
ggplotObject <- ecodata::plot_productivity_anomaly(report= 'NewEngland', varName= 'assessment' ,EPU= 'GB')
ggplotObject
```

```{r plot_productivity_anomalyNewEnglandassessmentGOM}
# Plot indicator
ggplotObject <- ecodata::plot_productivity_anomaly(report= 'NewEngland', varName= 'assessment' ,EPU= 'GOM')
ggplotObject
```


## Indicator statistics 
Spatial scale: _No response_

Temporal scale: _No response_

**Synthesis Theme**:

- [X] Multiple System Drivers


```{r autostats_productivity_anomaly}
# Either from Contributor or ecodata
```

## Implications
The apparent decline in productivity across multiple managed species in the MAB, along with mixed fish conditions in 2022, also suggest changing ecosystem productivity at multiple levels. During the 1990s high relative abundance of smaller bodied copepods and a lower relative abundance of *Calanus finmarchicus* was associated with regime shifts to higher fish recruitment [@perretti_regime_2017]. The unprecedented climate signals along with the trends toward lower productivity across multiple managed species indicate a need to continually evaluate whether management reference points remain appropriate, and to evaluate if ecosystem regime shifts have occurred or reorganization is in progress.

## Get the data

**Point of contact**: [Sarah.Gaichas@noaa.gov](mailto:Sarah.Gaichas@noaa.gov){.email}

**ecodata name**: `ecodata::productivity_anomaly`

**Variable definitions**

Variable names are organized using this format: [region] stock name - variable type and source  
Variables ending with "_Survey" are survey derived recruits/spawner anomalies 
Variables ending with "-Assessment" are assessment derived quantities Survey stock names are in ALL CAPS 
NE LME prepended to a survey stock name means the anomalies are coastwide 
Assessment stock names are in Sentence case  
Units for survey variables are the Z score of (number of recruits in year+1/biomass of adults in year) 
Units for plotted assessment variables are the Z score of (numbers of recruits per kg spawning biomass with recruits aligned to spawning biomass year using age at recruitment) 
Other variables are available in the assessment derived dataset but are not plotted. To be added later.

```{r vars_productivity_anomaly}
# Pull all var names
vars <- ecodata::productivity_anomaly |>
   dplyr::select(Var, Units) |>
   dplyr::distinct()

DT::datatable(vars)
```
**Indicator Category**:

- [X] Published Methods


## Public Availability

Source data are publicly available.

## Accessibility and Constraints

_No response_

**tech-doc link**
<https://noaa-edab.github.io/tech-doc/productivity_anomaly.html>

