# Recreational HMS {#rec_hms}

**Description**: Recreational shark landings pulled from the Marine Recreational Information Program (MRIP).

**Indicator family**: 

- [X] Megafauna


**Contributor(s)**: Brandon Beltz; Kim Bastille

**Affiliations**: NEFSC

```{r echo=FALSE}
knitr::opts_chunk$set(echo = F)
library(ecodata)
```
## Introduction to Indicator
Recreational shark landings pulled from the MRIP database

## Key Results and Visualizations
Recreational landings of sharks are plotted. Sharks are categorized as large coastal, pelagic, prohibited and small coastal.

### MAB

```{r plot_rec_hmsMAB}
# Plot indicator
ggplotObject <- ecodata::plot_rec_hms(report='MidAtlantic')
ggplotObject
```

### NE

```{r plot_rec_hmsNE}
# Plot indicator
ggplotObject <- ecodata::plot_rec_hms(report='NewEngland')
ggplotObject
```


## Indicator statistics 
Spatial scale: by EPU

Temporal scale: annually from 1981 to 2022

**Synthesis Theme**:

- [X] Multiple System Drivers


```{r autostats_rec_hms}
# Either from Contributor or ecodata
```

## Implications
Sharks are landed recreationally in quantities that are relevant to fisheries management. These landings should be considered when assessing the populations of sharks.

## Get the data

**Point of contact**: [Brandon Beltz (brandon.beltz@noaa.gov)](mailto:Brandon Beltz (brandon.beltz@noaa.gov)){.email}

**ecodata name**: `ecodata::rec_hms`

**Variable definitions**

See variables below

```{r vars_rec_hms}
# Pull all var names
vars <- ecodata::rec_hms |>
   dplyr::select(Var) |>
   dplyr::distinct()

DT::datatable(vars)
```
**Indicator Category**:

- [X] Database pull


## Public Availability

Source data are publicly available.

## Accessibility and Constraints

_No response_

**tech-doc link**
<https://noaa-edab.github.io/tech-doc/rec_hms.html>

