# Slopewater Proportions {#slopewater}

```{r echo=FALSE}
knitr::opts_chunk$set(echo = F)
library(ecodata)
library(downloadthis)
```

*Last updated July 23, 2026*

**Description**: This index gives the relative proportions of watermass type observed in the deep Northeast Channel (150-200 m water depth).

**Contributor(s)**: Paula Fratantoni

**Affiliations**: NEFSC

**Indicator Family**: 

- [X] Oceanographic


**Indicator Category**:

- [X] Extensive analysis, not yet published


**Synthesis Theme**:

- [X] Multiple System Drivers


```{r download_slopewater}
# Download dataset from ecodata
try(downloadthis::download_this(
ecodata::slopewater,
output_name = 'slopewater',
output_extension = '.csv',
button_label = 'Download dataset from `ecodata`',
button_type = 'default',
has_icon = TRUE,
icon = 'fa fa-save',
class = 'hvr-sweep-to-left',
csv2 = F),
silent = TRUE)
```

## Introduction to Indicator
Temperature and salinity measurements are examined to assess the composition of the waters entering the Gulf of Maine through the deep Northeast Channel. The analysis closely follows the methodology described by @mountain_labrador_2012. This method assumes that the waters flowing into the Northeast Channel between 150 and 200 meters depth are composed of slope waters, originating offshore of the continental shelf, and shelf waters, originating on the continental shelf south of Nova Scotia.

## Key Results and Visualizations
_No response_

**Mid-Atlantic**
```{r plot_slopewaterMidAtlantic}
# Plot indicator
ggplotObject <- ecodata::plot_slopewater(report = 'MidAtlantic', n = 10) +
ggplot2::ggtitle('')
ggplotObject
try(downloadthis::download_this(
ggplotObject$data,
output_name = 'slopewaterMidAtlantic',
output_extension = '.csv',
button_label = 'Download the data plotted above',
button_type = 'default',
has_icon = TRUE,
icon = 'fa fa-save',
class = 'hvr-sweep-to-left',
csv2 = F),
silent = TRUE)
```
<br>

**New England**
```{r plot_slopewaterNewEngland}
# Plot indicator
ggplotObject <- ecodata::plot_slopewater(report = 'NewEngland', n = 10) +
ggplot2::ggtitle('')
ggplotObject
try(downloadthis::download_this(
ggplotObject$data,
output_name = 'slopewaterNewEngland',
output_extension = '.csv',
button_label = 'Download the data plotted above',
button_type = 'default',
has_icon = TRUE,
icon = 'fa fa-save',
class = 'hvr-sweep-to-left',
csv2 = F),
silent = TRUE)
```
<br>

## Implications
_No response_

## Indicator statistics 
Spatial scale: Within the central Northeast Channel

Temporal scale: Annual

**Variable definitions**

```{r vars_slopewater}
# Pull all var names
vars <- ecodata::slopewater |>
   dplyr::select(Var, Units) |>
   dplyr::distinct()

DT::datatable(vars)
```
## Get the data

**Point of contact**: Paula Fratantoni (paula.fratantoni@noaa.gov)

**ecodata dataset**: `ecodata::slopewater`

**Tech Doc link**: <https://noaa-edab.github.io/tech-doc/slopewater.html>

### Public Availability

Source data are publicly available.

