# Slopewater Proportions {#slopewater}

**Description**: This index gives the relative proportions of watermass type observed in the deep Northeast Channel (150-200 m water depth).

**Indicator family**: 

- [X] Oceanographic


**Contributor(s)**: Paula Fratantoni

**Affiliations**: NEFSC

```{r echo=FALSE}
knitr::opts_chunk$set(echo = F)
library(ecodata)
```
## Introduction to Indicator
Temperature and salinity measurements are examined to assess the composition of the waters entering the Gulf of Maine through the deep Northeast Channel. The analysis closely follows the methodology described by @mountain_labrador_2012. This method assumes that the waters flowing into the Northeast Channel between 150 and 200 meters depth are composed of slope waters, originating offshore of the continental shelf, and shelf waters, originating on the continental shelf south of Nova Scotia.

## Key Results and Visualizations
_No response_

### NE

```{r plot_slopewaterNE}
# Plot indicator
ggplotObject <- ecodata::plot_slopewater(report='NewEngland')
ggplotObject
```


## Indicator statistics 
Spatial scale: Within the central Northeast Channel

Temporal scale: Annual

**Synthesis Theme**:

- [X] Multiple System Drivers


```{r autostats_slopewater}
# Either from Contributor or ecodata
```

## Implications
_No response_

## Get the data

**Point of contact**: [Paula Fratantoni (paula.fratantoni@noaa.gov)](mailto:Paula Fratantoni (paula.fratantoni@noaa.gov)){.email}

**ecodata name**: `ecodata::slopewater`

**Variable definitions**

Name: LSW; Definition: Percent total that is Labrador Slope Water; Units: percent; Name: WSW: Definition: Percent total that is Warm Slope Water; Units: percent

```{r vars_slopewater}
# Pull all var names
vars <- ecodata::slopewater |>
   dplyr::select(Var, Units) |>
   dplyr::distinct()

DT::datatable(vars)
```
**Indicator Category**:

- [X] Extensive analysis, not yet published


## Public Availability

Source data are publicly available.

## Accessibility and Constraints

_No response_

**tech-doc link**
<https://noaa-edab.github.io/tech-doc/slopewater.html>

