# Feeding guilds by management bodies {#species_groupings}

```{r echo=FALSE}
knitr::opts_chunk$set(echo = F)
library(ecodata)
library(downloadthis)
```

*Last updated July 23, 2026*

**Description**: Classification of species guild membership by management bodies.

**Contributor(s)**: Sarah Gaichas, Sean Lucey

**Affiliations**: NEFSC

**Indicator Family**: 

- [X] Megafauna
- [X] Economic


**Indicator Category**:

- [X] Database pull with analysis


**Synthesis Theme**:

- [x] Multiple System Drivers
- [x] Regime Shifts
- [x] Ecosystem Reorganization


```{r download_species_groupings}
# Download dataset from ecodata
try(downloadthis::download_this(
ecodata::species_groupings,
output_name = 'species_groupings',
output_extension = '.csv',
button_label = 'Download dataset from `ecodata`',
button_type = 'default',
has_icon = TRUE,
icon = 'fa fa-save',
class = 'hvr-sweep-to-left',
csv2 = F),
silent = TRUE)
```

## Introduction to Indicator
Feeding guilds are groups of species that feed similarly. At the ecosystem level, the food web is likely stable if overall biomass of feeding groups is stable over time, even if populations of species within the groups may be changing. 

We defined feeding guilds for fish and invertebrates captured by bottom trawl surveys using diet similarity, either from diet analysis or from literature [@garrison_dietary_2000; @smith_trophic_2010], and see [NEFSC food habits online](https://fwdp.shinyapps.io/tm2020/)

## Key Results and Visualizations
Each management body (MAFMC, NEFSC, State etc) has its managed species associated with different feeding guilds. This data set shows which managed species for each management body are in which guilds.

**Mid-Atlantic, 2024**
```{r plot_species_groupingsMidAtlantic2024}
# Plot indicator
ggplotObject <- ecodata::plot_species_groupings(report = 'MidAtlantic', varName = '2024') +
ggplot2::ggtitle('')
ggplotObject
try(downloadthis::download_this(
ggplotObject$data,
output_name = 'species_groupingsMidAtlantic2024',
output_extension = '.csv',
button_label = 'Download the data plotted above',
button_type = 'default',
has_icon = TRUE,
icon = 'fa fa-save',
class = 'hvr-sweep-to-left',
csv2 = F),
silent = TRUE)
```
<br>

**New England, 2024**
```{r plot_species_groupingsNewEngland2024}
# Plot indicator
ggplotObject <- ecodata::plot_species_groupings(report = 'NewEngland', varName = '2024') +
ggplot2::ggtitle('')
ggplotObject
try(downloadthis::download_this(
ggplotObject$data,
output_name = 'species_groupingsNewEngland2024',
output_extension = '.csv',
button_label = 'Download the data plotted above',
button_type = 'default',
has_icon = TRUE,
icon = 'fa fa-save',
class = 'hvr-sweep-to-left',
csv2 = F),
silent = TRUE)
```
<br>

## Implications
We changed species groupings in response to comments over the years. The table shows feeding guilds used in the 2024-2025 reports.  In 2025 an FMP column was added in response to Council requests.

## Indicator statistics 
Spatial scale: Coastwide

Temporal scale: N/A

**Variable definitions**

## Get the data

**Point of contact**: EDAB, edab.data@noaa.gov

**ecodata dataset**: `ecodata::species_groupings`

**Tech Doc link**: <https://noaa-edab.github.io/tech-doc/species_groupings.html>

### Public Availability

Source data are publicly available.

