# Stomach Fullness

**Description**: Stomach Fullness

**Contributor(s)**:  Laurel Smith
  
**Data steward**: Kimberly Bastille <kimberly.bastille@noaa.gov>
  
**Point of contact**: Laurel Smith <Laurel.smith@noaa.gov>

## Impact Summary

Description of why this indicator is important and what is going on in the Northeast related to this information. 

### Plotted data 

The plot below was built using the code found
[here](https://github.com/NOAA-EDAB/ecodata/blob/master/chunk-scripts/macrofauna.Rmd-ma-stomachs.R). 

```{r , code = readLines("https://raw.githubusercontent.com/NOAA-EDAB/ecodata/master/chunk-scripts/macrofauna.Rmd-ma-stomachs.R"), eval = T, fig.cap= "Stomach fullness anomaly."}

plotly::last_plot()
```

### Raw data 
 
```{r}
library(DT)
datatable(ecodata::stom_fullness, options = list(pageLength = 10))
```

## Methods

An index of stomach fullness was calculated from NEFSC autumn bottom trawl food habits data, as a simple ratio of estimated stomach content weight to total weight of an individual fish. Stomach fullness may be a better measure than absolute stomach weight if combining across species into a feeding guild, to prevent larger animals with heavier stomachs from dominating the index. An average stomach fullness was calcuated annually for each species and Ecological Production Unit (EPU).

### Data Sources

Stomach contents weights and individual fish weights (both to the nearest gram) were collected on the NEFSC bottom trawl surveys from 1992-present aboard RVs Albatross IV, Delaware II and the Henry B. Bigelow (see [Food Habits Database (FHDBS)](https://inport.nmfs.noaa.gov/inport) for access procedures).


### Data Extraction

NEFSC food habits data summarized in the R data file allfh.RData were obtained from Brian Smith (Brian.Smith@noaa.gov) for this index.


### Data Analysis

The stomach fullness index was calculated using the R script found [here](https://github.com/Laurels1/StomachFullness). 

### Data Processing

Fish stomach fullness index was formatted for inclusion in the `ecodata` R package using this [R code](https://github.com/NOAA-EDAB/ecodata/blob/master/data-raw/get_stom_fullness.R). Stomach fullness was expressed as an annual anomaly for each species in each region.


## References