# Sea Surface Temperature - Ocean Model {#surface_temp_mom6}

```{r echo=FALSE}
knitr::opts_chunk$set(echo = F)
library(ecodata)
library(downloadthis)
```

*Last updated July 23, 2026*

**Description**: The data presented here are seasonal sea surface temperature anomalies for the period 1993-2019 from the MOM6 hindcast ocean model for the Northwest Atlantic.  Also shown here are 10-year forecasts of sea surface temperature anomalies (2023-2033) from the MOM6 decadal forecast ocean model for the Northwest Atlantic.

**Contributor(s)**: Laura Gruenburg, Vimal Koul

**Affiliations**: NEFSC

**Indicator Family**: 

- [x] Oceanographic


**Indicator Category**:

- [x] Syntheses of published information
- [x] Database pull with analysis


**Synthesis Theme**:

- [x] Multiple System Drivers


```{r download_surface_temp_mom6}
# Download dataset from ecodata
try(downloadthis::download_this(
ecodata::surface_temp_mom6,
output_name = 'surface_temp_mom6',
output_extension = '.csv',
button_label = 'Download dataset from `ecodata`',
button_type = 'default',
has_icon = TRUE,
icon = 'fa fa-save',
class = 'hvr-sweep-to-left',
csv2 = F),
silent = TRUE)
```

## Introduction to Indicator
This indicator shows seasonal sea surface temperature anomalies from a regional ocean model for the Northwest Atlantic based on the Modular Ocean Model Version 6 (NWA MOM6).  The model is at roughly 1/12 a degree horizontal resolution with data from 1993-2019 from a hindcast simulation, which has no data assimilation.  Investigating the ability of this model to reproduce key indicators can help us to develop confidence in the model as well as learn where improvements can be made.

This indicator also shows 10-year forecasts of annual sea surface temperature anomalies from the decadal forecast of the MOM6 Northwest Atlantic regional ocean model. Understanding how temperatures might change in the future can help us anticipate potential shifts in marine communities.

## Key Results and Visualizations
Here we show seasonal sea surface temperature anomalies in the Mid-Atlantic Bight (MAB), Georges Bank (GB) and Gulf of Maine (GOM) EPUs.  Data from 1993 through 2019 are from NWA MOM6 (@ross_high-resolution_2023) with a base period for the climatology of 1993 through 2010.  We do not compare these anomalies to those from observed OISST as the base period for that dataset is 1982-2010.  

We also show the 10-year forecast for annual sea surface temperature anomalies in the MAB, GB, and GOM.  This forecast is from the decadal forecast of NWA MOM6 (@koul_predicted_2024), which was initialized in 2022 and ran for 10 years.  The forecast indicates a stagnation in surface temperature warming over the next ten years.  Retrospective forecasts, where a forecast is initiated in the past and output is compared to observations, have shown that the NWA MOM6 decadal forecast is able to correctly predict periods of temperature stability and rapid warming in this region (@koul_predicted_2024).

## Implications
Sea surface temperature is important for many marine species that have specific thermal preferences and tolerances.  When a region becomes too warm or cool, a species may respond in many ways including by shifting its range or changing its behavior.  Understanding how surface temperature may vary in the future can help us anticipate changes to marine communities and living marine resources.

## Indicator statistics 
Spatial scale: The model data is on a roughly 1/12 degree grid.  Here we show the data averaged by EPU.

Temporal scale: Winter (January, February, March), Spring (April, May, June), Summer (July, August, September), Fall (October, November, December), and Annual

**Variable definitions**

```{r vars_surface_temp_mom6}
# Pull all var names
vars <- ecodata::surface_temp_mom6 |>
   dplyr::select(Var, Units) |>
   dplyr::distinct()

DT::datatable(vars)
```
## Get the data

**Point of contact**: EDAB, edab.data@noaa.gov

**ecodata dataset**: `ecodata::surface_temp_mom6`

**Tech Doc link**: <https://noaa-edab.github.io/tech-doc/surface_temp_mom6.html>

### Public Availability

Source data are publicly available.

