# Bottom trawl survey diversity index {#survey_shannon}

**Description**: Diversity metric for the bottom trawl survey data

**Indicator family**: 

- [X] Megafauna


**Contributor(s)**: Andy Beet, Sean Lucey

**Affiliations**: NEFSC

```{r echo=FALSE}
knitr::opts_chunk$set(echo = F)
library(ecodata)
```
## Introduction to Indicator
Diversity metrics may be useful in determining shifts in species composition through time. The shannon-diversity index, a way to measure diversity using species richness and evenness is implemented in the attempt to determine changes over time. The index has a maximum value when all species are equally abundant.

## Key Results and Visualizations
A significant long term decline in the diversity metric is apparent in Georges Bank in both the Spring and the Fall. This is not seen in Gulf of Maine or the Mid Atlantic

```{r plot_survey_shannonMidAtlantic}
# Plot indicator
ggplotObject <- ecodata::plot_survey_shannon(report = 'MidAtlantic', n = 10)
ggplotObject
```

```{r plot_survey_shannonNewEngland}
# Plot indicator
ggplotObject <- ecodata::plot_survey_shannon(report = 'NewEngland', n = 10)
ggplotObject
```


## Indicator statistics 
Spatial scale: EPU

Temporal scale: SPRING, FALL

**Synthesis Theme**:

- [X] Regime Shifts
- [X] Ecosystem Reorganization


```{r autostats_survey_shannon}
# Either from Contributor or ecodata
```

## Implications
A decline in the shannon diversity index over time could be a result of several factors; a decline in the number of species being found in the system over time or a select few becoming more abundant (relative to the rest) over time. If the number of species remain contant through time, then the decline represents a shift to a more unbalanced, uneven system where some species are more abundannt that others. An increase would represent the opposite, a shift, in direction, to a more balanced, even system.

If the number of species is changing through time, the inference is more complicated since the maximum value of the shannon-index is a function of the number of species.

## Get the data

**Point of contact**: [andrew.beet@noaa.gov](mailto:andrew.beet@noaa.gov){.email}

**ecodata name**: `ecodata::survey_shannon`

**Variable definitions**

Name: NEFSC survey species diversity (SPRING, FALL) Units: Shannon index (unitless)

```{r vars_survey_shannon}
# Pull all var names
vars <- ecodata::survey_shannon |>
   dplyr::select(Var, Units) |>
   dplyr::distinct()

DT::datatable(vars)
```
**Indicator Category**:

- [X] Published Methods
- [X] Database pull


## Public Availability

Source data are publicly available.

## Accessibility and Constraints

_No response_

**tech-doc link**
<https://noaa-edab.github.io/tech-doc/survey_shannon.html>

