# Thermal Habitat Persistence

**Description**: The number of days per year per 1/12 degree cell that exceeds a temperature threshold.

**Indicator family**: 

- [X] Oceanographic
- [X] Habitat


**Contributor(s)**: Joe Caracappa

**Affiliations**: NEFSC

```{r echo=FALSE}
knitr::opts_chunk$set(echo = F)
library(ecodata)
```
## Introduction to Indicator
Many deep water benthic and demersal exhibit thermal preferences for metabolic, reproductive, and growth processes. Temperatures above these thermal presences may impair these processes. Data originate from GLORYS12V1 global reanalysis for 1993-01-01 to 2023-08-29, and from PSY forecasts from 2021-01-01 to 2023-12-31. Cells are mapped to 3 depth bins: 0-25m, 25-100m, and >100m. Two temperature thresholds are used representing a temperature where moderate (15C) and extreme (24C) thermal stresses are likely to occur across several species. GLORYS and PSY 1/12 degree grid is mapped to EPU_NOESTUARIES shape files by the center point of each grid cell.

## Key Results and Visualizations
Maps of full NE shelf with cells shading gradient showing number of days exceeding temperature thresholds. For 15C threshold, much of the southern MAB experienced >100 days, as well as along the shelf break and around Nantucket Shoals. In 2023, only the southern and/or shallow portion of the MAB exceeded the 24C threshold.


## Indicator statistics 
Spatial scale: full shelf

Temporal scale: annual

**Synthesis Theme**:

- [X] Multiple System Drivers


```{r autostats_thermal_habitat_persistence}
# Either from Contributor or ecodata
```

## Implications
Using a high resolution bottom temperature product allows for localized heat stress within EPUs that may be masked when looking at mean EPU bottom temperature. This can be important to sessile species who aren't able to move out of warm water conditions.

## Get the data

**Point of contact**: [joseph.caracappa@noaa.gov](mailto:joseph.caracappa@noaa.gov){.email}

**ecodata name**: `ecodata::thermal_habitat_persistence`

**Variable definitions**

Source: GLORYS (CMEM’s GLORYS12V1 global reanalysis bottom temperature) and PSY (CMEM’s PSY global forecast bottom temperature) 
min.depth: minimum of depth band max.depth: maximum of depth band 
temp.threshold: cutoff temperature for thermal area calculations (all areas greater than or equal to this temperature) 
longitude: longitude of cell center point latitude: latitude of cell center point 
Ndays: number of days exceeding temp.threshold

```{r vars_thermal_habitat_persistence}
# Pull all var names
vars <- ecodata::thermal_habitat_persistence |>
   dplyr::select(Var, Units) |>
   dplyr::distinct()

DT::datatable(vars)
```
**Indicator Category**:

- [X] Extensive analysis, not yet published


## Public Availability

Source data are publicly available.

## Accessibility and Constraints

_No response_

**tech-doc link**
<https://noaa-edab.github.io/tech-doc/thermal_habitat_persistence.html>

## References

