# Timing shifts: Risks to Seasonal Management

**Description**: Shifts in the timing of life-cycle events are a risk to meeting seasonal and temporal management objectives.

**Indicator family**: 

- [X] Oceanographic
- [X] Habitat
- [X] Lower trophic levels
- [X] Megafauna


**Contributor(s)**: Kimberly Hyde, Sarah Gaichas, Joe Carracappa

**Affiliations**: NEFSC

```{r echo=FALSE}
knitr::opts_chunk$set(echo = F)
library(ecodata)
```
## Introduction to Indicator
Changes in phenology, the seasonal timing of recurring life-cycle events, are a primary indicator of species responses to climate change [@staudinger_its_2019] (Staudinger et al., 2019). Observed phenological changes in the Northeast Shelf include spawning (REFS), migration [@crear_climate-influenced_2023] (Crear et al., 2023), prey availability, and ...  The phenological responses are often species-specific and vary depending on the primary environmental driver [@staudinger_its_2019].

## Key Results and Visualizations
TBD


## Indicator statistics 
Spatial scale: NES

Temporal scale: Seasonal

**Synthesis Theme**:

- [X] Multiple System Drivers


```{r autostats_timing_shifts}
# Either from Contributor or ecodata
```

## Implications
In progress

## Get the data

**Point of contact**: [nefsc.soe.leads@noaa.gov](mailto:nefsc.soe.leads@noaa.gov){.email}

**ecodata name**: No dataset

**Variable definitions**

NA


No Data

**Indicator Category**:

- [X] Syntheses of published information


## Public Availability

Source data are publicly available.

## Accessibility and Constraints

_No response_

**tech-doc link**
<https://noaa-edab.github.io/tech-doc/timing_shifts.html>

