# Transition Dates {#trans_dates}

```{r echo=FALSE}
knitr::opts_chunk$set(echo = F)
library(ecodata)
library(downloadthis)
```

*Last updated July 23, 2026*

**Description**: The date that cool winter conditions transition to warm stratified summer conditions.

**Contributor(s)**: Brandon Beltz, Kimberly Bastille, Kevin Friedland

**Affiliations**: NEFSC

**Indicator Family**: 

- [X] Oceanographic


**Indicator Category**:

- [X] Database pull


**Synthesis Theme**:

- [X] Multiple System Drivers


```{r download_trans_dates}
# Download dataset from ecodata
try(downloadthis::download_this(
ecodata::trans_dates,
output_name = 'trans_dates',
output_extension = '.csv',
button_label = 'Download dataset from `ecodata`',
button_type = 'default',
has_icon = TRUE,
icon = 'fa fa-save',
class = 'hvr-sweep-to-left',
csv2 = F),
silent = TRUE)
```

## Introduction to Indicator
Transition dates are defined as the day of the year when surface temperatures changeover from cool to warm conditions in the spring and back to cool conditions in the fall.

## Key Results and Visualizations
Ocean summer length in Mid-Atlantic: the annual total number of days between the spring thermal transition date and the fall thermal transition date. The transition dates are defined as the day of the year when surface temperatures changeover from cool to warm conditions in the spring and back to cool conditions in the fall.

**timing, Mid-Atlantic**
```{r plot_trans_datestimingMidAtlantic}
# Plot indicator
ggplotObject <- ecodata::plot_trans_dates(varName = 'timing', report = 'MidAtlantic', n = 10) +
ggplot2::ggtitle('')
ggplotObject
try(downloadthis::download_this(
ggplotObject$data,
output_name = 'trans_datestimingMidAtlantic',
output_extension = '.csv',
button_label = 'Download the data plotted above',
button_type = 'default',
has_icon = TRUE,
icon = 'fa fa-save',
class = 'hvr-sweep-to-left',
csv2 = F),
silent = TRUE)
```
<br>

**length, Mid-Atlantic**
```{r plot_trans_dateslengthMidAtlantic}
# Plot indicator
ggplotObject <- ecodata::plot_trans_dates(varName = 'length', report = 'MidAtlantic', n = 10) +
ggplot2::ggtitle('')
ggplotObject
try(downloadthis::download_this(
ggplotObject$data,
output_name = 'trans_dateslengthMidAtlantic',
output_extension = '.csv',
button_label = 'Download the data plotted above',
button_type = 'default',
has_icon = TRUE,
icon = 'fa fa-save',
class = 'hvr-sweep-to-left',
csv2 = F),
silent = TRUE)
```
<br>

**timing, New England**
```{r plot_trans_datestimingNewEngland}
# Plot indicator
ggplotObject <- ecodata::plot_trans_dates(varName = 'timing', report = 'NewEngland', n = 10) +
ggplot2::ggtitle('')
ggplotObject
try(downloadthis::download_this(
ggplotObject$data,
output_name = 'trans_datestimingNewEngland',
output_extension = '.csv',
button_label = 'Download the data plotted above',
button_type = 'default',
has_icon = TRUE,
icon = 'fa fa-save',
class = 'hvr-sweep-to-left',
csv2 = F),
silent = TRUE)
```
<br>

**length, New England**
```{r plot_trans_dateslengthNewEngland}
# Plot indicator
ggplotObject <- ecodata::plot_trans_dates(varName = 'length', report = 'NewEngland', n = 10) +
ggplot2::ggtitle('')
ggplotObject
try(downloadthis::download_this(
ggplotObject$data,
output_name = 'trans_dateslengthNewEngland',
output_extension = '.csv',
button_label = 'Download the data plotted above',
button_type = 'default',
has_icon = TRUE,
icon = 'fa fa-save',
class = 'hvr-sweep-to-left',
csv2 = F),
silent = TRUE)
```
<br>

## Implications
Prolonged fall temperatures have been linked to the increased number of cold-stunned Kemp’s ridley sea turtles found in Cape Cod Bay @griffin_warming_2019

## Indicator statistics 
Spatial scale: by EPU

Temporal scale: Annual time series (1982 to present)

**Variable definitions**

```{r vars_trans_dates}
# Pull all var names
vars <- ecodata::trans_dates |>
   dplyr::select(Var) |>
   dplyr::distinct()

DT::datatable(vars)
```
## Get the data

**Point of contact**: EDAB, edab.data@noaa.gov

**ecodata dataset**: `ecodata::trans_dates`

**Tech Doc link**: <https://noaa-edab.github.io/tech-doc/trans_dates.html>

### Public Availability

Source data are publicly available.

