# Transition Dates {#trans_dates}

**Description**: The date that cool winter conditions transition to warm stratified summer conditions.

**Indicator family**: 

- [X] Oceanographic


**Contributor(s)**: Brandon Beltz, Kimberly Bastille, Kevin Friedland

**Affiliations**: NEFSC

```{r echo=FALSE}
knitr::opts_chunk$set(echo = F)
library(ecodata)
```
## Introduction to Indicator
Transition dates are defined as the day of the year when surface temperatures changeover from cool to warm conditions in the spring and back to cool conditions in the fall.

## Key Results and Visualizations
Ocean summer length in Mid-Atlantic: the annual total number of days between the spring thermal transition date and the fall thermal transition date. The transition dates are defined as the day of the year when surface temperatures changeover from cool to warm conditions in the spring and back to cool conditions in the fall.

```{r plot_trans_datestimingMidAtlantic}
# Plot indicator
ggplotObject <- ecodata::plot_trans_dates(varName = 'timing', report = 'MidAtlantic', n = 10)
ggplotObject
```

```{r plot_trans_dateslengthMidAtlantic}
# Plot indicator
ggplotObject <- ecodata::plot_trans_dates(varName = 'length', report = 'MidAtlantic', n = 10)
ggplotObject
```

```{r plot_trans_datestimingNewEngland}
# Plot indicator
ggplotObject <- ecodata::plot_trans_dates(varName = 'timing', report = 'NewEngland', n = 10)
ggplotObject
```

```{r plot_trans_dateslengthNewEngland}
# Plot indicator
ggplotObject <- ecodata::plot_trans_dates(varName = 'length', report = 'NewEngland', n = 10)
ggplotObject
```


## Indicator statistics 
Spatial scale: by EPU

Temporal scale: Annual time series (1982 to present)

**Synthesis Theme**:

- [X] Multiple System Drivers


```{r autostats_trans_dates}
# Either from Contributor or ecodata
```

## Implications
Prolonged fall temperatures have been linked to the increased number of cold-stunned Kemp’s ridley sea turtles found in Cape Cod Bay @griffin_warming_2019

## Get the data

**Point of contact**: [EDAB, edab.data@noaa.gov](mailto:EDAB, edab.data@noaa.gov){.email}

**ecodata name**: `ecodata::trans_dates`

**Variable definitions**

1. falltrans: fall transition day 2. sprtrans: spring transition day 3. maxday: max sst day

```{r vars_trans_dates}
# Pull all var names
vars <- ecodata::trans_dates |>
   dplyr::select(Var) |>
   dplyr::distinct()

DT::datatable(vars)
```
**Indicator Category**:

- [X] Database pull


## Public Availability

Source data are publicly available.

## Accessibility and Constraints

_No response_

**tech-doc link**
<https://noaa-edab.github.io/tech-doc/trans_dates.html>

