# Community Port Landings and Revenue from Wind Energy Areas (WEAs)

**Description**: NA

**Indicator family**: 

- [X] Social
- [X] Economic


**Contributor(s)**: Angela Silva, Doug Christel

**Affiliations**: ?

```{r echo=FALSE}
knitr::opts_chunk$set(echo = F)
library(ecodata)
```
## Introduction to Indicator
NA

## Key Results and Visualizations
This figure links historic port revenue (2008-2022) from within all wind energy areas (including all lease areas, the Central Atlantic proposed lease areas A-2, B-1 and C-1, GOM Draft Wind Energy Area and secondary areas), as a proportion of a port’s total fisheries revenue based on vessel trip reports as described in the revenue and landings of species in the wind indicator above. The range (minimum and maximum) of total percent fisheries revenue from within wind energy areas is presented in the graph and ports are sorted from greatest to least fisheries revenue from within wind areas. Those communities that score Med-High or higher in at least one of the vulnerability indicators that address environmental justice concerns (i.e., Poverty, Population Composition, Personal Disruption; see indicator definitions) are noted with a triangle. Gentrification pressure is also highlighted here, with those communities that score Med-High or higher in one or more gentrification pressure indicators (i.e., Housing Disruption, Retiree Migration, Urban Sprawl) represented with a circle.

### MAB

```{r plot_wind_portMAB}
# Plot indicator
ggplotObject <- ecodata::plot_wind_port(report='MidAtlantic')
ggplotObject
```

### NE

```{r plot_wind_portNE}
# Plot indicator
ggplotObject <- ecodata::plot_wind_port(report='NewEngland')
ggplotObject
```


## Indicator statistics 
Spatial scale: Full Shelf, broken down into Mid-Atlantic and New England communities

Temporal scale: 2008-2022

**Synthesis Theme**:




```{r autostats_wind_port}
# Either from Contributor or ecodata
```

## Implications
NA

## Get the data

**Point of contact**: [Angela Silva (angela.silva@noaa.gov)](mailto:Angela Silva (angela.silva@noaa.gov)){.email}

**ecodata name**: `ecodata::wind_port`

**Variable definitions**

NA

```{r vars_wind_port}
# Pull all var names
vars <- ecodata::wind_port |>
   dplyr::select(Var) |>
   dplyr::distinct()

DT::datatable(vars)
```
**Indicator Category**:

- [X] Database pull with analysis


## Public Availability

Source data are publicly available.

## Accessibility and Constraints

_No response_

**tech-doc link**
<https://noaa-edab.github.io/tech-doc/wind_port.html>

