# Community Port Landings and Revenue from Wind Energy Areas (WEAs)

**Description**: NA

**Indicator family**: 

- [X] Social
- [X] Economic


**Contributor(s)**: Angela Silva, Doug Christel

**Affiliations**: ?

```{r echo=FALSE}
knitr::opts_chunk$set(echo = F)
library(ecodata)
```
## Introduction to Indicator
NA

## Key Results and Visualizations
NA

### MAB

```{r plot_wind_portMAB}
# Plot indicator
ggplotObject <- ecodata::plot_wind_port(report='MidAtlantic')
print(ggplotObject)
```

### NE

```{r plot_wind_portNE}
# Plot indicator
ggplotObject <- ecodata::plot_wind_port(report='NewEngland')
print(ggplotObject)
```


## Indicator statistics 
Spatial scale: Full Shelf, broken down into Mid-Atlantic and New England communities

Temporal scale: 2008-2022

**Synthesis Theme**:




```{r autostats_wind_port}
# Either from Contributor or ecodata
```

## Implications
NA

## Get the data

**Point of contact**: [Angela Silva (angela.silva@noaa.gov)](mailto:Angela Silva (angela.silva@noaa.gov)){.email}

**ecodata name**: `ecodata::wind_port`

**Variable definitions**

NA

```{r vars_wind_port}
# Pull all var names
vars <- ecodata::wind_port |>
   dplyr::select(Var) |>
   dplyr::distinct()

DT::datatable(vars)
```
**Indicator Category**:

- [X] Database pull with analysis


## Public Availability

Source data are publicly available.

## Accessibility and Constraints

_No response_

**tech-doc link**
<https://noaa-edab.github.io/tech-doc/wind_port.html>

## References

