# Zooplankton Abundance Anomalies {#zoo_abundance_anom}

**Description**: Abundance anomalies for 20 zooplankton taxa

**Indicator family**: 

- [X] Lower trophic levels


**Contributor(s)**: Harvey Walsh, Ryan Morse, Kevin Friedland, Mike Jones

**Affiliations**: NEFSC

```{r echo=FALSE}
knitr::opts_chunk$set(echo = F)
library(ecodata)
```
## Introduction to Indicator
Zooplankton represent a critical trophic link from primary producers to fish in marine ecosystems.

## Key Results and Visualizations
Abundance anomalies of small and large copepods have varied over time by EPU.
Small bodied copepods and cnidarians show increasing trends in all EPUs. 
Large bodied copepods and euphausiids show no significant trend in any EPU.

```{r plot_zoo_abundance_anomcopepodMidAtlantic}
# Plot indicator
ggplotObject <- ecodata::plot_zoo_abundance_anom(varName = 'copepod', report = 'MidAtlantic', n = 10)
ggplotObject
```

```{r plot_zoo_abundance_anomeuphausidMidAtlantic}
# Plot indicator
ggplotObject <- ecodata::plot_zoo_abundance_anom(varName = 'euphausid', report = 'MidAtlantic', n = 10)
ggplotObject
```

```{r plot_zoo_abundance_anomcopepodNewEngland}
# Plot indicator
ggplotObject <- ecodata::plot_zoo_abundance_anom(varName = 'copepod', report = 'NewEngland', n = 10)
ggplotObject
```

```{r plot_zoo_abundance_anomeuphausidNewEngland}
# Plot indicator
ggplotObject <- ecodata::plot_zoo_abundance_anom(varName = 'euphausid', report = 'NewEngland', n = 10)
ggplotObject
```


## Indicator statistics 
Spatial scale: by EPU

Temporal scale: Annual

**Synthesis Theme**:

- [X] Multiple System Drivers


```{r autostats_zoo_abundance_anom}
# Either from Contributor or ecodata
```

## Implications
Check these. If they are correct, we could be seeing the less energy dense zooplankton becoming more abundant in each system.

## Get the data

**Point of contact**: [Harvey Walsh, harvey.walsh@noaa.gov; Ryan Morse, ryan.morse@noaa.gov](mailto:Harvey Walsh, harvey.walsh@noaa.gov; Ryan Morse, ryan.morse@noaa.gov){.email}

**ecodata name**: `ecodata::zoo_abundance_anom`

**Variable definitions**

All are unitless anomalies from the 1977-2020 mean abundance for each taxon. Variables are taxa names:  (to be described by contributors) "Calfin"         "LgCopepods"     "SmCopepods"     
"Cyclopoida"     "Diplostraca"    "Ostracoda"      "Cirripedia"     "Euphausiacea"  "Gammaridea"     "Hyperiidea"     "Mysidacea"      "Decapoda"       "Polychaeta"     "Echinodermata"  
"Mollusca"     "Pteropod"      "Chaetognatha"   "Cnidaria"       "Tunicate"       "Protozoa"

```{r vars_zoo_abundance_anom}
# Pull all var names
vars <- ecodata::zoo_abundance_anom |>
   dplyr::select(Var, Units) |>
   dplyr::distinct()

DT::datatable(vars)
```
**Indicator Category**:

- [X] Published Methods
- [X] Extensive analysis, not yet published
- [X] Syntheses of published information
- [X] Database pull with analysis


## Public Availability

Source data are NOT publicly available.

## Accessibility and Constraints

Request from Harvey Walsh, harvey.walsh@noaa.gov

**tech-doc link**
<https://noaa-edab.github.io/tech-doc/zoo_abundance_anom.html>

