# Zooplankton Abundance Anomalies {#zoo_abundance_anom}

```{r echo=FALSE}
knitr::opts_chunk$set(echo = F)
library(ecodata)
library(downloadthis)
```

*Last updated July 23, 2026*

**Description**: Abundance anomalies for 20 zooplankton taxa

**Contributor(s)**: Harvey Walsh, Ryan Morse, Kevin Friedland, Mike Jones

**Affiliations**: NEFSC

**Indicator Family**: 

- [X] Lower trophic levels


**Indicator Category**:

- [X] Published Methods
- [X] Extensive analysis, not yet published
- [X] Syntheses of published information
- [X] Database pull with analysis


**Synthesis Theme**:

- [X] Multiple System Drivers


```{r download_zoo_abundance_anom}
# Download dataset from ecodata
try(downloadthis::download_this(
ecodata::zoo_abundance_anom,
output_name = 'zoo_abundance_anom',
output_extension = '.csv',
button_label = 'Download dataset from `ecodata`',
button_type = 'default',
has_icon = TRUE,
icon = 'fa fa-save',
class = 'hvr-sweep-to-left',
csv2 = F),
silent = TRUE)
```

## Introduction to Indicator
Zooplankton represent a critical trophic link from primary producers to fish in marine ecosystems.

## Key Results and Visualizations
Abundance anomalies of small and large copepods have varied over time by EPU.
Small bodied copepods and cnidarians show increasing trends in all EPUs. 
Large bodied copepods and euphausiids show no significant trend in any EPU.

**copepod, Mid-Atlantic**
```{r plot_zoo_abundance_anomcopepodMidAtlantic}
# Plot indicator
ggplotObject <- ecodata::plot_zoo_abundance_anom(varName = 'copepod', report = 'MidAtlantic', n = 10) +
ggplot2::ggtitle('')
ggplotObject
try(downloadthis::download_this(
ggplotObject$data,
output_name = 'zoo_abundance_anomcopepodMidAtlantic',
output_extension = '.csv',
button_label = 'Download the data plotted above',
button_type = 'default',
has_icon = TRUE,
icon = 'fa fa-save',
class = 'hvr-sweep-to-left',
csv2 = F),
silent = TRUE)
```
<br>

**euphausid, Mid-Atlantic**
```{r plot_zoo_abundance_anomeuphausidMidAtlantic}
# Plot indicator
ggplotObject <- ecodata::plot_zoo_abundance_anom(varName = 'euphausid', report = 'MidAtlantic', n = 10) +
ggplot2::ggtitle('')
ggplotObject
try(downloadthis::download_this(
ggplotObject$data,
output_name = 'zoo_abundance_anomeuphausidMidAtlantic',
output_extension = '.csv',
button_label = 'Download the data plotted above',
button_type = 'default',
has_icon = TRUE,
icon = 'fa fa-save',
class = 'hvr-sweep-to-left',
csv2 = F),
silent = TRUE)
```
<br>

**copepod, New England**
```{r plot_zoo_abundance_anomcopepodNewEngland}
# Plot indicator
ggplotObject <- ecodata::plot_zoo_abundance_anom(varName = 'copepod', report = 'NewEngland', n = 10) +
ggplot2::ggtitle('')
ggplotObject
try(downloadthis::download_this(
ggplotObject$data,
output_name = 'zoo_abundance_anomcopepodNewEngland',
output_extension = '.csv',
button_label = 'Download the data plotted above',
button_type = 'default',
has_icon = TRUE,
icon = 'fa fa-save',
class = 'hvr-sweep-to-left',
csv2 = F),
silent = TRUE)
```
<br>

**euphausid, New England**
```{r plot_zoo_abundance_anomeuphausidNewEngland}
# Plot indicator
ggplotObject <- ecodata::plot_zoo_abundance_anom(varName = 'euphausid', report = 'NewEngland', n = 10) +
ggplot2::ggtitle('')
ggplotObject
try(downloadthis::download_this(
ggplotObject$data,
output_name = 'zoo_abundance_anomeuphausidNewEngland',
output_extension = '.csv',
button_label = 'Download the data plotted above',
button_type = 'default',
has_icon = TRUE,
icon = 'fa fa-save',
class = 'hvr-sweep-to-left',
csv2 = F),
silent = TRUE)
```
<br>

## Implications
Check these. If they are correct, we could be seeing the less energy dense zooplankton becoming more abundant in each system.

## Indicator statistics 
Spatial scale: by EPU

Temporal scale: Annual

**Variable definitions**

```{r vars_zoo_abundance_anom}
# Pull all var names
vars <- ecodata::zoo_abundance_anom |>
   dplyr::select(Var, Units) |>
   dplyr::distinct()

DT::datatable(vars)
```
## Get the data

**Point of contact**: Harvey Walsh (harvey.walsh@noaa.gov)

**ecodata dataset**: `ecodata::zoo_abundance_anom`

**Tech Doc link**: <https://noaa-edab.github.io/tech-doc/zoo_abundance_anom.html>

### Public Availability

Source data are NOT publicly available.

### Accessibility Constraints

Request from Harvey Walsh, harvey.walsh@noaa.gov

