# Zooplankton Indices {#zooplankton_index}

**Description**: Model-based abundance and annual center of gravity indices for zooplankton groups sampled by NEFSC surveys

**Indicator family**: 

- [X] Lower trophic levels


**Contributor(s)**: Abigail Tyrell, Sarah Gaichas, Harvey Walsh

**Affiliations**: NEFSC

```{r echo=FALSE}
knitr::opts_chunk$set(echo = F)
library(ecodata)
```
## Introduction to Indicator
Zooplankton are primary consumers in marine ecosystems that transfer energy from phytoplankton to fish, marine mammals, and birds. The Northeast Fisheries Science Center has conducted [zooplankton surveys](https://www.fisheries.noaa.gov/inport/item/9286) since the 1970s. Spatially explicit indices of abundance for several zooplankton groups were estimated using spatio-temporal modeling (VAST, @thorson_comparing_2017; @thorson_guidance_2019). 

The zooplankton groups in the model include: 

**Copepods**

* _Calanus finmarchicus_, = Single species group, used to represent large copepods in the [small-large index](https://noaa-edab.github.io/tech-doc/zoo_abundance_anom.html))

* Large copepods ALL: _Calanus finmarchicus_, _Metridia lucens_, _Calanus minor_, _Eucalanus_ spp., _Calanus_ spp.

* Small copepods ALL: _Centropages typicus_, _Pseudocalanus_ spp., T_emora longicornis_, _Centropages hamatus_, _Paracalanus parvus_, _Acartia_ spp., _Clausocalanus arcuicornis_, _Acartia longiremis_, _Clausocalanus furcatus_, _Temora stylifera_, _Temora_ spp., _Tortanus discaudatus_, _Paracalanus_ spp.

* Small copeopods SOE (used to represent small copepods in the small-large index): _Centropages typicus_, _Pseudocalanus_ spp., _Temora longicornis_, _Centropages hamatus_

**Euphausiids** (krill) modeled as family level group; not differentiated to species in data

**Zooplankton volume** a metric of total zooplankton; all groups combined displacement volume in the collection net

## Key Results and Visualizations
**Abundance**
Large and small copepod groups are mainly fluctuating without trend in both spring and fall across the EPUs. However, a significant long term decrease in small copepods was apparent in the MAB during fall, and a recent decrease during spring. Similarly, zooplankton volume is fluctuating without long term trends over time. A recent significant decline was detected for Spring zooplankton volume in the MAB. 

Euphausiids, in contrast, show increasing trends over time in all regions during fall, and in the MAB during spring. In addition, a significant recent increase was detected in the GOM in Fall. 


**Center of Gravity**
Small copepods in fall are trending northeast similar to forage fish and aggregated survey species. Small copepods in spring are trending north. Large copepods in fall are trending west, similar to benthos. No significant change for Calfin, Euphausiids, or Zooplankton volume center of gravity over time.

```{r plot_zooplankton_indexMidAtlanticCalfinindex}
# Plot indicator
ggplotObject <- ecodata::plot_zooplankton_index(report = 'MidAtlantic', varName = 'Calfin', plottype = 'index', n = 10)
ggplotObject
```

```{r plot_zooplankton_indexNewEnglandCalfinindex}
# Plot indicator
ggplotObject <- ecodata::plot_zooplankton_index(report = 'NewEngland', varName = 'Calfin', plottype = 'index', n = 10)
ggplotObject
```

```{r plot_zooplankton_indexMidAtlanticEuphindex}
# Plot indicator
ggplotObject <- ecodata::plot_zooplankton_index(report = 'MidAtlantic', varName = 'Euph', plottype = 'index', n = 10)
ggplotObject
```

```{r plot_zooplankton_indexNewEnglandEuphindex}
# Plot indicator
ggplotObject <- ecodata::plot_zooplankton_index(report = 'NewEngland', varName = 'Euph', plottype = 'index', n = 10)
ggplotObject
```

```{r plot_zooplankton_indexMidAtlanticSmallcopesoeindex}
# Plot indicator
ggplotObject <- ecodata::plot_zooplankton_index(report = 'MidAtlantic', varName = 'Smallcopesoe', plottype = 'index', n = 10)
ggplotObject
```

```{r plot_zooplankton_indexNewEnglandSmallcopesoeindex}
# Plot indicator
ggplotObject <- ecodata::plot_zooplankton_index(report = 'NewEngland', varName = 'Smallcopesoe', plottype = 'index', n = 10)
ggplotObject
```

```{r plot_zooplankton_indexMidAtlanticLgcopeallindex}
# Plot indicator
ggplotObject <- ecodata::plot_zooplankton_index(report = 'MidAtlantic', varName = 'Lgcopeall', plottype = 'index', n = 10)
ggplotObject
```

```{r plot_zooplankton_indexNewEnglandLgcopeallindex}
# Plot indicator
ggplotObject <- ecodata::plot_zooplankton_index(report = 'NewEngland', varName = 'Lgcopeall', plottype = 'index', n = 10)
ggplotObject
```

```{r plot_zooplankton_indexMidAtlanticZoopvolindex}
# Plot indicator
ggplotObject <- ecodata::plot_zooplankton_index(report = 'MidAtlantic', varName = 'Zoopvol', plottype = 'index', n = 10)
ggplotObject
```

```{r plot_zooplankton_indexNewEnglandZoopvolindex}
# Plot indicator
ggplotObject <- ecodata::plot_zooplankton_index(report = 'NewEngland', varName = 'Zoopvol', plottype = 'index', n = 10)
ggplotObject
```

```{r plot_zooplankton_indexMidAtlanticSmallcopeallindex}
# Plot indicator
ggplotObject <- ecodata::plot_zooplankton_index(report = 'MidAtlantic', varName = 'Smallcopeall', plottype = 'index', n = 10)
ggplotObject
```

```{r plot_zooplankton_indexNewEnglandSmallcopeallindex}
# Plot indicator
ggplotObject <- ecodata::plot_zooplankton_index(report = 'NewEngland', varName = 'Smallcopeall', plottype = 'index', n = 10)
ggplotObject
```

```{r plot_zooplankton_indexMidAtlanticCalfincog}
# Plot indicator
ggplotObject <- ecodata::plot_zooplankton_index(report = 'MidAtlantic', varName = 'Calfin', plottype = 'cog', n = 10)
ggplotObject
```

```{r plot_zooplankton_indexNewEnglandCalfincog}
# Plot indicator
ggplotObject <- ecodata::plot_zooplankton_index(report = 'NewEngland', varName = 'Calfin', plottype = 'cog', n = 10)
ggplotObject
```

```{r plot_zooplankton_indexMidAtlanticEuphcog}
# Plot indicator
ggplotObject <- ecodata::plot_zooplankton_index(report = 'MidAtlantic', varName = 'Euph', plottype = 'cog', n = 10)
ggplotObject
```

```{r plot_zooplankton_indexNewEnglandEuphcog}
# Plot indicator
ggplotObject <- ecodata::plot_zooplankton_index(report = 'NewEngland', varName = 'Euph', plottype = 'cog', n = 10)
ggplotObject
```

```{r plot_zooplankton_indexMidAtlanticSmallcopesoecog}
# Plot indicator
ggplotObject <- ecodata::plot_zooplankton_index(report = 'MidAtlantic', varName = 'Smallcopesoe', plottype = 'cog', n = 10)
ggplotObject
```

```{r plot_zooplankton_indexNewEnglandSmallcopesoecog}
# Plot indicator
ggplotObject <- ecodata::plot_zooplankton_index(report = 'NewEngland', varName = 'Smallcopesoe', plottype = 'cog', n = 10)
ggplotObject
```

```{r plot_zooplankton_indexMidAtlanticLgcopeallcog}
# Plot indicator
ggplotObject <- ecodata::plot_zooplankton_index(report = 'MidAtlantic', varName = 'Lgcopeall', plottype = 'cog', n = 10)
ggplotObject
```

```{r plot_zooplankton_indexNewEnglandLgcopeallcog}
# Plot indicator
ggplotObject <- ecodata::plot_zooplankton_index(report = 'NewEngland', varName = 'Lgcopeall', plottype = 'cog', n = 10)
ggplotObject
```

```{r plot_zooplankton_indexMidAtlanticZoopvolcog}
# Plot indicator
ggplotObject <- ecodata::plot_zooplankton_index(report = 'MidAtlantic', varName = 'Zoopvol', plottype = 'cog', n = 10)
ggplotObject
```

```{r plot_zooplankton_indexNewEnglandZoopvolcog}
# Plot indicator
ggplotObject <- ecodata::plot_zooplankton_index(report = 'NewEngland', varName = 'Zoopvol', plottype = 'cog', n = 10)
ggplotObject
```

```{r plot_zooplankton_indexMidAtlanticSmallcopeallcog}
# Plot indicator
ggplotObject <- ecodata::plot_zooplankton_index(report = 'MidAtlantic', varName = 'Smallcopeall', plottype = 'cog', n = 10)
ggplotObject
```

```{r plot_zooplankton_indexNewEnglandSmallcopeallcog}
# Plot indicator
ggplotObject <- ecodata::plot_zooplankton_index(report = 'NewEngland', varName = 'Smallcopeall', plottype = 'cog', n = 10)
ggplotObject
```


## Indicator statistics 
Spatial scale: by EPU and full shelf

Temporal scale: Spring (January-June), Fall (July-December)

**Synthesis Theme**:

- [X] Multiple System Drivers


```{r autostats_zooplankton_index}
# Either from Contributor or ecodata
```

## Implications
Fluctuations in zooplankton community composition can affect both the consumption of phytoplankton vs the export of energy to the benthos, as well as the availability of prey for zooplankton feeding forage fish, marine mammals, and seabirds. 

The relative dominance of large and small copepods in Northeast US ecosystems has been liked to changes in fish productivity  [@perretti_regime_2017; morse_distinct_2017].

Distribution shifts for zooplankton may result in mismatches with zooplankton feeding fish, marine mammal, and seabirds.

## Get the data

**Point of contact**: [EDAB, edab.data@noaa.gov](mailto:EDAB, edab.data@noaa.gov){.email}

**ecodata name**: `ecodata::zooplankton_index`

**Variable definitions**

Naming key:  1) *Calanus finmarchicus* (calfin_100m3) = “calfin” 2) Large copepods (calfin_100m3, mlucens_100m3, calminor_100m3, euc_100m3, calspp_100m3) = “lgcopeALL” 
3) Small copepods (all) (ctyp_100m3, pseudo_100m3, tlong_100m3, cham_100m3, para_100m3, acarspp_100m3, clauso, acarlong_100m3, fur_100m3, ost_100m3, temspp_100m3, tort_100m3, paraspp_100m3) = “smallcopeALL” 
4) Small copepods (SOE) (ctyp_100m3, pseudo_100m3, tlong_100m3, cham_100m3) = “smallcopeSOE”

```{r vars_zooplankton_index}
# Pull all var names
vars <- ecodata::zooplankton_index |>
   dplyr::select(Var, Units) |>
   dplyr::distinct()

DT::datatable(vars)
```
**Indicator Category**:

- [X] Extensive analysis, not yet published


## Public Availability

Source data are NOT publicly available.

## Accessibility and Constraints

Request from Harvey Walsh, harvey.walsh@noaa.gov

**tech-doc link**
<https://noaa-edab.github.io/tech-doc/zooplankton_index.html>

