# Zooplankton Indices {#zooplankton_index}

```{r echo=FALSE}
knitr::opts_chunk$set(echo = F)
library(ecodata)
library(downloadthis)
```

*Last updated July 23, 2026*

**Description**: Model-based abundance and annual center of gravity indices for zooplankton groups sampled by NEFSC surveys

**Contributor(s)**: Abigail Tyrell, Sarah Gaichas, Harvey Walsh

**Affiliations**: NEFSC

**Indicator Family**: 

- [X] Lower trophic levels


**Indicator Category**:

- [X] Extensive analysis, not yet published


**Synthesis Theme**:

- [X] Multiple System Drivers


```{r download_zooplankton_index}
# Download dataset from ecodata
try(downloadthis::download_this(
ecodata::zooplankton_index,
output_name = 'zooplankton_index',
output_extension = '.csv',
button_label = 'Download dataset from `ecodata`',
button_type = 'default',
has_icon = TRUE,
icon = 'fa fa-save',
class = 'hvr-sweep-to-left',
csv2 = F),
silent = TRUE)
```

## Introduction to Indicator
Zooplankton are primary consumers in marine ecosystems that transfer energy from phytoplankton to fish, marine mammals, and birds. The Northeast Fisheries Science Center has conducted [zooplankton surveys](https://www.fisheries.noaa.gov/inport/item/9286) since the 1970s. Spatially explicit indices of abundance for several zooplankton groups were estimated using spatio-temporal modeling (VAST, @thorson_comparing_2017; @thorson_guidance_2019). 

The zooplankton groups in the model include: 

**Copepods**

* _Calanus finmarchicus_, = Single species group, used to represent large copepods in the [small-large index](https://noaa-edab.github.io/tech-doc/zoo_abundance_anom.html))

* Large copepods ALL: _Calanus finmarchicus_, _Metridia lucens_, _Calanus minor_, _Eucalanus_ spp., _Calanus_ spp.

* Small copepods ALL: _Centropages typicus_, _Pseudocalanus_ spp., T_emora longicornis_, _Centropages hamatus_, _Paracalanus parvus_, _Acartia_ spp., _Clausocalanus arcuicornis_, _Acartia longiremis_, _Clausocalanus furcatus_, _Temora stylifera_, _Temora_ spp., _Tortanus discaudatus_, _Paracalanus_ spp.

* Small copeopods SOE (used to represent small copepods in the small-large index): _Centropages typicus_, _Pseudocalanus_ spp., _Temora longicornis_, _Centropages hamatus_

**Euphausiids** (krill) modeled as family level group; not differentiated to species in data

**Zooplankton volume** a metric of total zooplankton; all groups combined displacement volume in the collection net

## Key Results and Visualizations
**Abundance** Large and small copepod groups are mainly fluctuating without trend in both spring and fall across the EPUs. Large copepod biomass has a short-term decline in the spring in both the GOM and GB. Small copepods have a long-term decline in the MAB in fall, and a long-term increase in the GOM in the spring. Euphausiid show increasing trends over time in all regions during fall. Zooplankton volume is fluctuating without long term trends over time.

**Center of Gravity** Small copepods in fall are trending northeast similar to forage fish and aggregated survey species. Small copepods in spring are trending north. There is a short-term westward trend for Calanus finmarchicus in the spring. There has been no other significant change for Calanus finmarchicus, large copepods, euphausiids, or zooplankton volume center of gravity over time.

**Mid-Atlantic, Calfin, index**
```{r plot_zooplankton_indexMidAtlanticCalfinindex}
# Plot indicator
ggplotObject <- ecodata::plot_zooplankton_index(report = 'MidAtlantic', varName = 'Calfin', plottype = 'index', n = 10) +
ggplot2::ggtitle('')
ggplotObject
try(downloadthis::download_this(
ggplotObject$data,
output_name = 'zooplankton_indexMidAtlanticCalfinindex',
output_extension = '.csv',
button_label = 'Download the data plotted above',
button_type = 'default',
has_icon = TRUE,
icon = 'fa fa-save',
class = 'hvr-sweep-to-left',
csv2 = F),
silent = TRUE)
```
<br>

**New England, Calfin, index**
```{r plot_zooplankton_indexNewEnglandCalfinindex}
# Plot indicator
ggplotObject <- ecodata::plot_zooplankton_index(report = 'NewEngland', varName = 'Calfin', plottype = 'index', n = 10) +
ggplot2::ggtitle('')
ggplotObject
try(downloadthis::download_this(
ggplotObject$data,
output_name = 'zooplankton_indexNewEnglandCalfinindex',
output_extension = '.csv',
button_label = 'Download the data plotted above',
button_type = 'default',
has_icon = TRUE,
icon = 'fa fa-save',
class = 'hvr-sweep-to-left',
csv2 = F),
silent = TRUE)
```
<br>

**Mid-Atlantic, Euph, index**
```{r plot_zooplankton_indexMidAtlanticEuphindex}
# Plot indicator
ggplotObject <- ecodata::plot_zooplankton_index(report = 'MidAtlantic', varName = 'Euph', plottype = 'index', n = 10) +
ggplot2::ggtitle('')
ggplotObject
try(downloadthis::download_this(
ggplotObject$data,
output_name = 'zooplankton_indexMidAtlanticEuphindex',
output_extension = '.csv',
button_label = 'Download the data plotted above',
button_type = 'default',
has_icon = TRUE,
icon = 'fa fa-save',
class = 'hvr-sweep-to-left',
csv2 = F),
silent = TRUE)
```
<br>

**New England, Euph, index**
```{r plot_zooplankton_indexNewEnglandEuphindex}
# Plot indicator
ggplotObject <- ecodata::plot_zooplankton_index(report = 'NewEngland', varName = 'Euph', plottype = 'index', n = 10) +
ggplot2::ggtitle('')
ggplotObject
try(downloadthis::download_this(
ggplotObject$data,
output_name = 'zooplankton_indexNewEnglandEuphindex',
output_extension = '.csv',
button_label = 'Download the data plotted above',
button_type = 'default',
has_icon = TRUE,
icon = 'fa fa-save',
class = 'hvr-sweep-to-left',
csv2 = F),
silent = TRUE)
```
<br>

**Mid-Atlantic, Smallcopesoe, index**
```{r plot_zooplankton_indexMidAtlanticSmallcopesoeindex}
# Plot indicator
ggplotObject <- ecodata::plot_zooplankton_index(report = 'MidAtlantic', varName = 'Smallcopesoe', plottype = 'index', n = 10) +
ggplot2::ggtitle('')
ggplotObject
try(downloadthis::download_this(
ggplotObject$data,
output_name = 'zooplankton_indexMidAtlanticSmallcopesoeindex',
output_extension = '.csv',
button_label = 'Download the data plotted above',
button_type = 'default',
has_icon = TRUE,
icon = 'fa fa-save',
class = 'hvr-sweep-to-left',
csv2 = F),
silent = TRUE)
```
<br>

**New England, Smallcopesoe, index**
```{r plot_zooplankton_indexNewEnglandSmallcopesoeindex}
# Plot indicator
ggplotObject <- ecodata::plot_zooplankton_index(report = 'NewEngland', varName = 'Smallcopesoe', plottype = 'index', n = 10) +
ggplot2::ggtitle('')
ggplotObject
try(downloadthis::download_this(
ggplotObject$data,
output_name = 'zooplankton_indexNewEnglandSmallcopesoeindex',
output_extension = '.csv',
button_label = 'Download the data plotted above',
button_type = 'default',
has_icon = TRUE,
icon = 'fa fa-save',
class = 'hvr-sweep-to-left',
csv2 = F),
silent = TRUE)
```
<br>

**Mid-Atlantic, Lgcopeall, index**
```{r plot_zooplankton_indexMidAtlanticLgcopeallindex}
# Plot indicator
ggplotObject <- ecodata::plot_zooplankton_index(report = 'MidAtlantic', varName = 'Lgcopeall', plottype = 'index', n = 10) +
ggplot2::ggtitle('')
ggplotObject
try(downloadthis::download_this(
ggplotObject$data,
output_name = 'zooplankton_indexMidAtlanticLgcopeallindex',
output_extension = '.csv',
button_label = 'Download the data plotted above',
button_type = 'default',
has_icon = TRUE,
icon = 'fa fa-save',
class = 'hvr-sweep-to-left',
csv2 = F),
silent = TRUE)
```
<br>

**New England, Lgcopeall, index**
```{r plot_zooplankton_indexNewEnglandLgcopeallindex}
# Plot indicator
ggplotObject <- ecodata::plot_zooplankton_index(report = 'NewEngland', varName = 'Lgcopeall', plottype = 'index', n = 10) +
ggplot2::ggtitle('')
ggplotObject
try(downloadthis::download_this(
ggplotObject$data,
output_name = 'zooplankton_indexNewEnglandLgcopeallindex',
output_extension = '.csv',
button_label = 'Download the data plotted above',
button_type = 'default',
has_icon = TRUE,
icon = 'fa fa-save',
class = 'hvr-sweep-to-left',
csv2 = F),
silent = TRUE)
```
<br>

**Mid-Atlantic, Zoopvol, index**
```{r plot_zooplankton_indexMidAtlanticZoopvolindex}
# Plot indicator
ggplotObject <- ecodata::plot_zooplankton_index(report = 'MidAtlantic', varName = 'Zoopvol', plottype = 'index', n = 10) +
ggplot2::ggtitle('')
ggplotObject
try(downloadthis::download_this(
ggplotObject$data,
output_name = 'zooplankton_indexMidAtlanticZoopvolindex',
output_extension = '.csv',
button_label = 'Download the data plotted above',
button_type = 'default',
has_icon = TRUE,
icon = 'fa fa-save',
class = 'hvr-sweep-to-left',
csv2 = F),
silent = TRUE)
```
<br>

**New England, Zoopvol, index**
```{r plot_zooplankton_indexNewEnglandZoopvolindex}
# Plot indicator
ggplotObject <- ecodata::plot_zooplankton_index(report = 'NewEngland', varName = 'Zoopvol', plottype = 'index', n = 10) +
ggplot2::ggtitle('')
ggplotObject
try(downloadthis::download_this(
ggplotObject$data,
output_name = 'zooplankton_indexNewEnglandZoopvolindex',
output_extension = '.csv',
button_label = 'Download the data plotted above',
button_type = 'default',
has_icon = TRUE,
icon = 'fa fa-save',
class = 'hvr-sweep-to-left',
csv2 = F),
silent = TRUE)
```
<br>

**Mid-Atlantic, Smallcopeall, index**
```{r plot_zooplankton_indexMidAtlanticSmallcopeallindex}
# Plot indicator
ggplotObject <- ecodata::plot_zooplankton_index(report = 'MidAtlantic', varName = 'Smallcopeall', plottype = 'index', n = 10) +
ggplot2::ggtitle('')
ggplotObject
try(downloadthis::download_this(
ggplotObject$data,
output_name = 'zooplankton_indexMidAtlanticSmallcopeallindex',
output_extension = '.csv',
button_label = 'Download the data plotted above',
button_type = 'default',
has_icon = TRUE,
icon = 'fa fa-save',
class = 'hvr-sweep-to-left',
csv2 = F),
silent = TRUE)
```
<br>

**New England, Smallcopeall, index**
```{r plot_zooplankton_indexNewEnglandSmallcopeallindex}
# Plot indicator
ggplotObject <- ecodata::plot_zooplankton_index(report = 'NewEngland', varName = 'Smallcopeall', plottype = 'index', n = 10) +
ggplot2::ggtitle('')
ggplotObject
try(downloadthis::download_this(
ggplotObject$data,
output_name = 'zooplankton_indexNewEnglandSmallcopeallindex',
output_extension = '.csv',
button_label = 'Download the data plotted above',
button_type = 'default',
has_icon = TRUE,
icon = 'fa fa-save',
class = 'hvr-sweep-to-left',
csv2 = F),
silent = TRUE)
```
<br>

**Mid-Atlantic, Calfin, cog**
```{r plot_zooplankton_indexMidAtlanticCalfincog}
# Plot indicator
ggplotObject <- ecodata::plot_zooplankton_index(report = 'MidAtlantic', varName = 'Calfin', plottype = 'cog', n = 10) +
ggplot2::ggtitle('')
ggplotObject
try(downloadthis::download_this(
ggplotObject$data,
output_name = 'zooplankton_indexMidAtlanticCalfincog',
output_extension = '.csv',
button_label = 'Download the data plotted above',
button_type = 'default',
has_icon = TRUE,
icon = 'fa fa-save',
class = 'hvr-sweep-to-left',
csv2 = F),
silent = TRUE)
```
<br>

**New England, Calfin, cog**
```{r plot_zooplankton_indexNewEnglandCalfincog}
# Plot indicator
ggplotObject <- ecodata::plot_zooplankton_index(report = 'NewEngland', varName = 'Calfin', plottype = 'cog', n = 10) +
ggplot2::ggtitle('')
ggplotObject
try(downloadthis::download_this(
ggplotObject$data,
output_name = 'zooplankton_indexNewEnglandCalfincog',
output_extension = '.csv',
button_label = 'Download the data plotted above',
button_type = 'default',
has_icon = TRUE,
icon = 'fa fa-save',
class = 'hvr-sweep-to-left',
csv2 = F),
silent = TRUE)
```
<br>

**Mid-Atlantic, Euph, cog**
```{r plot_zooplankton_indexMidAtlanticEuphcog}
# Plot indicator
ggplotObject <- ecodata::plot_zooplankton_index(report = 'MidAtlantic', varName = 'Euph', plottype = 'cog', n = 10) +
ggplot2::ggtitle('')
ggplotObject
try(downloadthis::download_this(
ggplotObject$data,
output_name = 'zooplankton_indexMidAtlanticEuphcog',
output_extension = '.csv',
button_label = 'Download the data plotted above',
button_type = 'default',
has_icon = TRUE,
icon = 'fa fa-save',
class = 'hvr-sweep-to-left',
csv2 = F),
silent = TRUE)
```
<br>

**New England, Euph, cog**
```{r plot_zooplankton_indexNewEnglandEuphcog}
# Plot indicator
ggplotObject <- ecodata::plot_zooplankton_index(report = 'NewEngland', varName = 'Euph', plottype = 'cog', n = 10) +
ggplot2::ggtitle('')
ggplotObject
try(downloadthis::download_this(
ggplotObject$data,
output_name = 'zooplankton_indexNewEnglandEuphcog',
output_extension = '.csv',
button_label = 'Download the data plotted above',
button_type = 'default',
has_icon = TRUE,
icon = 'fa fa-save',
class = 'hvr-sweep-to-left',
csv2 = F),
silent = TRUE)
```
<br>

**Mid-Atlantic, Smallcopesoe, cog**
```{r plot_zooplankton_indexMidAtlanticSmallcopesoecog}
# Plot indicator
ggplotObject <- ecodata::plot_zooplankton_index(report = 'MidAtlantic', varName = 'Smallcopesoe', plottype = 'cog', n = 10) +
ggplot2::ggtitle('')
ggplotObject
try(downloadthis::download_this(
ggplotObject$data,
output_name = 'zooplankton_indexMidAtlanticSmallcopesoecog',
output_extension = '.csv',
button_label = 'Download the data plotted above',
button_type = 'default',
has_icon = TRUE,
icon = 'fa fa-save',
class = 'hvr-sweep-to-left',
csv2 = F),
silent = TRUE)
```
<br>

**New England, Smallcopesoe, cog**
```{r plot_zooplankton_indexNewEnglandSmallcopesoecog}
# Plot indicator
ggplotObject <- ecodata::plot_zooplankton_index(report = 'NewEngland', varName = 'Smallcopesoe', plottype = 'cog', n = 10) +
ggplot2::ggtitle('')
ggplotObject
try(downloadthis::download_this(
ggplotObject$data,
output_name = 'zooplankton_indexNewEnglandSmallcopesoecog',
output_extension = '.csv',
button_label = 'Download the data plotted above',
button_type = 'default',
has_icon = TRUE,
icon = 'fa fa-save',
class = 'hvr-sweep-to-left',
csv2 = F),
silent = TRUE)
```
<br>

**Mid-Atlantic, Lgcopeall, cog**
```{r plot_zooplankton_indexMidAtlanticLgcopeallcog}
# Plot indicator
ggplotObject <- ecodata::plot_zooplankton_index(report = 'MidAtlantic', varName = 'Lgcopeall', plottype = 'cog', n = 10) +
ggplot2::ggtitle('')
ggplotObject
try(downloadthis::download_this(
ggplotObject$data,
output_name = 'zooplankton_indexMidAtlanticLgcopeallcog',
output_extension = '.csv',
button_label = 'Download the data plotted above',
button_type = 'default',
has_icon = TRUE,
icon = 'fa fa-save',
class = 'hvr-sweep-to-left',
csv2 = F),
silent = TRUE)
```
<br>

**New England, Lgcopeall, cog**
```{r plot_zooplankton_indexNewEnglandLgcopeallcog}
# Plot indicator
ggplotObject <- ecodata::plot_zooplankton_index(report = 'NewEngland', varName = 'Lgcopeall', plottype = 'cog', n = 10) +
ggplot2::ggtitle('')
ggplotObject
try(downloadthis::download_this(
ggplotObject$data,
output_name = 'zooplankton_indexNewEnglandLgcopeallcog',
output_extension = '.csv',
button_label = 'Download the data plotted above',
button_type = 'default',
has_icon = TRUE,
icon = 'fa fa-save',
class = 'hvr-sweep-to-left',
csv2 = F),
silent = TRUE)
```
<br>

**Mid-Atlantic, Zoopvol, cog**
```{r plot_zooplankton_indexMidAtlanticZoopvolcog}
# Plot indicator
ggplotObject <- ecodata::plot_zooplankton_index(report = 'MidAtlantic', varName = 'Zoopvol', plottype = 'cog', n = 10) +
ggplot2::ggtitle('')
ggplotObject
try(downloadthis::download_this(
ggplotObject$data,
output_name = 'zooplankton_indexMidAtlanticZoopvolcog',
output_extension = '.csv',
button_label = 'Download the data plotted above',
button_type = 'default',
has_icon = TRUE,
icon = 'fa fa-save',
class = 'hvr-sweep-to-left',
csv2 = F),
silent = TRUE)
```
<br>

**New England, Zoopvol, cog**
```{r plot_zooplankton_indexNewEnglandZoopvolcog}
# Plot indicator
ggplotObject <- ecodata::plot_zooplankton_index(report = 'NewEngland', varName = 'Zoopvol', plottype = 'cog', n = 10) +
ggplot2::ggtitle('')
ggplotObject
try(downloadthis::download_this(
ggplotObject$data,
output_name = 'zooplankton_indexNewEnglandZoopvolcog',
output_extension = '.csv',
button_label = 'Download the data plotted above',
button_type = 'default',
has_icon = TRUE,
icon = 'fa fa-save',
class = 'hvr-sweep-to-left',
csv2 = F),
silent = TRUE)
```
<br>

**Mid-Atlantic, Smallcopeall, cog**
```{r plot_zooplankton_indexMidAtlanticSmallcopeallcog}
# Plot indicator
ggplotObject <- ecodata::plot_zooplankton_index(report = 'MidAtlantic', varName = 'Smallcopeall', plottype = 'cog', n = 10) +
ggplot2::ggtitle('')
ggplotObject
try(downloadthis::download_this(
ggplotObject$data,
output_name = 'zooplankton_indexMidAtlanticSmallcopeallcog',
output_extension = '.csv',
button_label = 'Download the data plotted above',
button_type = 'default',
has_icon = TRUE,
icon = 'fa fa-save',
class = 'hvr-sweep-to-left',
csv2 = F),
silent = TRUE)
```
<br>

**New England, Smallcopeall, cog**
```{r plot_zooplankton_indexNewEnglandSmallcopeallcog}
# Plot indicator
ggplotObject <- ecodata::plot_zooplankton_index(report = 'NewEngland', varName = 'Smallcopeall', plottype = 'cog', n = 10) +
ggplot2::ggtitle('')
ggplotObject
try(downloadthis::download_this(
ggplotObject$data,
output_name = 'zooplankton_indexNewEnglandSmallcopeallcog',
output_extension = '.csv',
button_label = 'Download the data plotted above',
button_type = 'default',
has_icon = TRUE,
icon = 'fa fa-save',
class = 'hvr-sweep-to-left',
csv2 = F),
silent = TRUE)
```
<br>

## Implications
Fluctuations in zooplankton community composition can affect both the consumption of phytoplankton vs the export of energy to the benthos, as well as the availability of prey for zooplankton feeding forage fish, marine mammals, and seabirds. 

The relative dominance of large and small copepods in Northeast US ecosystems has been liked to changes in fish productivity  [@perretti_regime_2017].

Distribution shifts for zooplankton may result in mismatches with zooplankton feeding fish, marine mammal, and seabirds.

## Indicator statistics 
Spatial scale: by EPU and full shelf

Temporal scale: Spring (January-June), Fall (July-December)

**Variable definitions**

```{r vars_zooplankton_index}
# Pull all var names
vars <- ecodata::zooplankton_index |>
   dplyr::select(Var, Units) |>
   dplyr::distinct()

DT::datatable(vars)
```
## Get the data

**Point of contact**: EDAB, edab.data@noaa.gov

**ecodata dataset**: `ecodata::zooplankton_index`

**Tech Doc link**: <https://noaa-edab.github.io/tech-doc/zooplankton_index.html>

### Public Availability

Source data are NOT publicly available.

### Accessibility Constraints

Request from Harvey Walsh, harvey.walsh@noaa.gov

